---
title: "Factorial Experiment Visualization (v2)"
author: "Anthony Stewart"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
    fig-width: 16
    fig-height: 9
execute:
  warning: false
  message: false
---

Walkthrough of the v2 factorial experiment results (8 configs x 5 seeds x
{multiclass, binary}) plus the follow-on studies (patch-count learning curve,
UNet3+ comparison). Inputs are the small CSV/JSON aggregation outputs only:

- `dl_08_aggregate_factorial.py` -> `Models/factorial_results_v2/<mode>/analysis/`
- `dl_08b_aggregate_patchcurve.py` -> `Models/results_patchcurve_v2/<mode>/analysis/` and `Models/results_arch_v2/<mode>/analysis/`
- `dl_09b_shap_geomorph_export.py` -> `Models/factorial_results_v2/<mode>/analysis/shap_geomorph_channels.csv`

```{r}
#| label: setup

library(tidyverse)
library(patchwork)
library(jsonlite)
library(colorspace)
library(scales)
```


```{r}
#| label: factorial-data

multi_dir  <- "../Models/factorial_results_v2/multiclass"
binary_dir <- "../Models/factorial_results_v2/binary"

multi_fact_data <- read.csv(file.path(multi_dir, "analysis/factorial_long.csv"))
glimpse(multi_fact_data)

binary_fact_data <- read.csv(file.path(binary_dir, "analysis/factorial_long.csv"))
glimpse(binary_fact_data)
```

```{r}
#| label: data-organizing

config_order <- c("fld_nolidar_leafon", "fld_nolidar_leafoff",  "fld_chmret_leafon", "fld_chmret_leafoff",
                 "nwi_chmret_leafoff", "nwiextra_chmret_leafoff", "nwifield_chmret_leafoff", "flddeg_chmret_leafoff")

config_label <- c("fld_nolidar_leafon" = "Field Data\n No Lidar\n No Leaf-Off",
                  "fld_nolidar_leafoff" = "Field Data\n No Lidar\n With Leaf-Off",
                  "fld_chmret_leafon" = "Field Data\n With Lidar\n No Leaf-Off",
                  "fld_chmret_leafoff" = "Field Data\n With Lidar\n With Leaf-Off",
                  "nwi_chmret_leafoff" = "NWI Data\n With Lidar\n With Leaf-Off",
                  "nwiextra_chmret_leafoff" = "NWI x2 Data\n With Lidar\n With Leaf-Off",
                  "nwifield_chmret_leafoff" = "NWI + Field\n With Lidar\n With Leaf-Off",
                  "flddeg_chmret_leafoff" = "Degraded Field\n With Lidar\n With Leaf-Off")

# single-line variant for heatmap y-axes (stacked labels overlap there)
config_label_1line <- gsub("\n ", ", ", config_label)

class_order <- c("EMW", "FSW", "SSW", "WET", "UPL")
class_color <- c("FSW" = "#81b14f", "SSW" = "#c4bd8b", "EMW" = "#8faab3", "UPL" = "#6a4836", "WET" = "#388C73")

metric_label <- c("precision" = "Precision", "recall" = "Recall", "f1" = "F1", "iou" = "IoU",
                  "macro_f1" = "Macro F1", "mean_iou" = "Mean IoU", "overall_accuracy" = "Overall accuracy")

organize <- function(df) {
  df |>
    mutate(config = factor(config, levels = config_order),
           class  = factor(class,  levels = class_order))
}
multi_fact_data  <- organize(multi_fact_data)
binary_fact_data <- organize(binary_fact_data)

# Both modes in one frame for faceted figures
fact_data <- bind_rows(
  multi_fact_data  |> mutate(mode = "Multiclass"),
  binary_fact_data |> mutate(mode = "Binary")
) |> mutate(mode = factor(mode, levels = c("Multiclass", "Binary")))

theme_param <- theme(
  text = element_text(size = 18),
  legend.position = 'right',
  legend.text = element_text(size = rel(0.70), hjust = 0.5),
  panel.background = element_blank(),
  panel.grid.major = element_line(colour = "grey50", linewidth = 0.2),
  axis.ticks = element_blank(),
  axis.text = element_text(size = rel(0.9)),
  axis.title = element_text(size = rel(1)),
  strip.background = element_blank(),
  strip.text = element_text(size = rel(0.9), face = "bold")
)

theme_heatmap <- theme_param + theme(panel.grid.major = element_blank())

# Boxplot + jittered seed points, shared across the metric figures
box_jitter <- function(point_size = 5) {
  list(
    geom_boxplot(outlier.shape = NA, position = position_dodge(width = 0.75)),
    geom_jitter(position = position_jitterdodge(jitter.width = 0.15, dodge.width = 0.75),
                size = point_size, shape = 21, colour = "grey15")
  )
}

# Single-hue sequential fill for the heatmaps (magnitude = light -> dark)
seq_fill <- function(...) {
  scale_fill_gradient(low = "#f2f7f5", high = "#14523f", na.value = "grey93", ...)
}
```


### Does leaf off imagery or lidar-derived metrics improve field-based models?

Structure

- Just using the the field-verified dataset models
- Model config on the x-axis, model evaluation metric on the y-axis
- box and whiskers plots representing the model measure spread, points for the individual model measures
- Precision and recall share a panel: full class color = precision, lighter shade = recall

```{r}
#| label: q1-precision-recall
#| fig-width: 20
#| fig-height: 9

# fill palette: "<class> Precision" = class color, "<class> Recall" = lighter shade
pr_fill <- c(
  setNames(class_color, paste(names(class_color), "Precision")),
  setNames(lighten(class_color, 0.45), paste(names(class_color), "Recall"))
)
pr_levels <- paste(rep(class_order, each = 2), c("Precision", "Recall"))

pr_data <- fact_data |>
  filter(label == "fld") |>
  pivot_longer(c(precision, recall), names_to = "metric", values_to = "value") |>
  mutate(class_metric = factor(paste(class, metric_label[metric]), levels = pr_levels))

pr_plot <- function(which_mode) {
  pr_data |>
    filter(mode == which_mode) |>
    ggplot(aes(x = config, y = value, fill = class_metric)) +
      box_jitter(point_size = 3.5) +
      scale_fill_manual(values = pr_fill, name = NULL) +
      scale_x_discrete(labels = config_label) +
      guides(fill = guide_legend(ncol = 1)) +
      labs(x = NULL, y = "Precision / Recall", title = which_mode) +
      theme_param
}

pr_plot("Multiclass") + pr_plot("Binary")
```

```{r}
#| label: q1-f1-iou
#| fig-width: 18
#| fig-height: 12

fact_data |>
  filter(label == "fld") |>
  pivot_longer(c(f1, iou), names_to = "metric", values_to = "value") |>
  mutate(metric = factor(metric_label[metric], levels = c("F1", "IoU"))) |>
  ggplot(aes(x = config, y = value, fill = class)) +
    box_jitter() +
    facet_grid(metric ~ mode) +
    scale_fill_manual(values = class_color) +
    scale_x_discrete(labels = config_label) +
    labs(x = NULL, y = NULL) +
    theme_param
```


### How do NWI data labels compare to field and GIS-derived data?

#### Also does degrading the field data hinder its performance?

- All label sources on the full feature set (`chmret_leafoff`), evaluated against field labels

```{r}
#| label: q2-nwi-f1-iou
#| fig-width: 20
#| fig-height: 12

fact_data |>
  filter(featureset == "chmret_leafoff") |>
  pivot_longer(c(f1, iou), names_to = "metric", values_to = "value") |>
  mutate(metric = factor(metric_label[metric], levels = c("F1", "IoU"))) |>
  ggplot(aes(x = config, y = value, fill = class)) +
    box_jitter() +
    facet_grid(metric ~ mode) +
    scale_fill_manual(values = class_color) +
    scale_x_discrete(labels = config_label) +
    labs(x = NULL, y = NULL) +
    theme_param
```

Model-level summary (macro metrics are repeated on every class row of
`factorial_long.csv`, so collapse to one row per model first):

```{r}
#| label: q2-macro
#| fig-width: 20
#| fig-height: 8

fact_data |>
  filter(featureset == "chmret_leafoff") |>
  distinct(mode, config, seed, macro_f1, mean_iou) |>
  pivot_longer(c(macro_f1, mean_iou), names_to = "metric", values_to = "value") |>
  mutate(metric = metric_label[metric]) |>
  ggplot(aes(x = config, y = value, fill = metric)) +
    box_jitter(point_size = 4) +
    facet_grid(. ~ mode) +
    scale_fill_manual(values = c("Macro F1" = "#388C73", "Mean IoU" = lighten("#388C73", 0.45)), name = NULL) +
    scale_x_discrete(labels = config_label) +
    labs(x = NULL, y = NULL) +
    theme_param
```


### How do field and NWI models compare across each other

- Heat maps of model metrics: config (rows) x class (cols), one panel per metric.

```{r}
#| label: q3-metric-heatmap
#| fig-width: 18
#| fig-height: 14

metric_heat <- fact_data |>
  pivot_longer(c(precision, recall, f1, iou), names_to = "metric", values_to = "value") |>
  mutate(metric = factor(metric_label[metric], levels = c("Precision", "Recall", "F1", "IoU"))) |>
  group_by(mode, config, class, metric) |>
  summarise(value = mean(value), .groups = "drop")

heat_panel <- function(which_mode) {
  metric_heat |>
    filter(mode == which_mode) |>
    ggplot(aes(x = class, y = fct_rev(config), fill = value)) +
      geom_tile(colour = "white", linewidth = 1) +
      geom_text(aes(label = number(value, accuracy = 0.01),
                    colour = value > 0.55), size = 4.5, show.legend = FALSE) +
      facet_grid(. ~ metric) +
      scale_colour_manual(values = c(`TRUE` = "white", `FALSE` = "grey20")) +
      seq_fill(limits = c(0, 1), name = "Seed-mean\nscore") +
      scale_y_discrete(labels = config_label_1line) +
      labs(x = NULL, y = NULL, title = which_mode) +
      theme_heatmap
}

heat_panel("Multiclass") / heat_panel("Binary") +
  plot_layout(guides = "collect", heights = c(1, 1))
```


### How accurate are the classes via confusion matrices?

- binary and multiclass model confusion matrices for each configuration
- color and fill by relative % but include a parentheses with the number of pixels e.g. 70% (3,400,000)
- rows (true class) sum to 100%, so the diagonal reads as recall; counts are seed-mean pixels

```{r}
#| label: q4-confusion-load

read_confusion <- function(mode_dir) {
  map_dfr(config_order, function(cfg) {
    f <- file.path(mode_dir, "analysis/confusion_mean", paste0(cfg, ".csv"))
    if (!file.exists(f)) return(NULL)
    read.csv(f, row.names = 1, check.names = FALSE) |>
      rownames_to_column("true") |>
      pivot_longer(-true, names_to = "pred", values_to = "n") |>
      mutate(config = factor(cfg, levels = config_order))
  }) |>
    group_by(config, true) |>
    mutate(prop = n / sum(n)) |>
    ungroup() |>
    mutate(true = factor(true, levels = rev(class_order)),
           pred = factor(pred, levels = class_order),
           cell_label = paste0(percent(prop, accuracy = 1), "\n(",
                               number(n, accuracy = 0.1, scale_cut = cut_short_scale()), ")"))
}

confusion_plot <- function(mode_dir, title, text_size = 4.5) {
  read_confusion(mode_dir) |>
    ggplot(aes(x = pred, y = true, fill = prop)) +
      geom_tile(colour = "white", linewidth = 1) +
      geom_text(aes(label = cell_label, colour = prop > 0.55),
                size = text_size, lineheight = 0.9, show.legend = FALSE) +
      facet_wrap(~ config, ncol = 4, labeller = labeller(config = config_label)) +
      scale_colour_manual(values = c(`TRUE` = "white", `FALSE` = "grey20")) +
      seq_fill(limits = c(0, 1), labels = percent, name = "% of\ntrue class") +
      labs(x = "Predicted", y = "True", title = title) +
      theme_heatmap
}
```

```{r}
#| label: q4-confusion-multi
#| fig-width: 20
#| fig-height: 11

confusion_plot(multi_dir, "Multiclass")
```

```{r}
#| label: q4-confusion-binary
#| fig-width: 20
#| fig-height: 8

confusion_plot(binary_dir, "Binary", text_size = 5.5)
```


### What were the important features in each model configuration according to SHAP?

- Model heatmap with cross-model comparison: each model normalized to its own total importance, so bands are comparable across configs. Bands a config doesn't use are blank.
- Uses the PER-CHANNEL MEAN aggregation (not the raw sum) so the one-hot Geomorph_local band is not inflated 10x relative to single-channel bands.
- Also show within the Geomorphons category a heatmap of the different categories across models

```{r}
#| label: q5-shap-load

band_order <- c("DEM", "slope_local", "Geomorph_local", "flowacc", "twi",
                "CHM", "pct_below_1m", "pct_1m_to_5m", "pct_above_5m",
                "r", "g", "b", "nir", "r_lo", "g_lo", "b_lo", "nir_lo")

band_label <- c("DEM" = "DEM", "slope_local" = "Slope", "Geomorph_local" = "Geomorphons",
                "flowacc" = "Flow accum.", "twi" = "TWI",
                "CHM" = "CHM", "pct_below_1m" = "% returns < 1 m",
                "pct_1m_to_5m" = "% returns 1-5 m", "pct_above_5m" = "% returns > 5 m",
                "r" = "Red (leaf-on)", "g" = "Green (leaf-on)", "b" = "Blue (leaf-on)", "nir" = "NIR (leaf-on)",
                "r_lo" = "Red (leaf-off)", "g_lo" = "Green (leaf-off)", "b_lo" = "Blue (leaf-off)", "nir_lo" = "NIR (leaf-off)")

read_shap_mode <- function(mode_dir) {
  files <- Sys.glob(file.path(mode_dir, "*", "seed*", "shap", "*_shap_importance.json"))
  map_dfr(files, function(f) {
    j <- fromJSON(f)
    tibble(
      config = basename(dirname(dirname(dirname(f)))),
      seed   = parse_number(str_extract(f, "seed\\d+")),
      band   = names(j$importance_overall_per_channel),
      value  = unlist(j$importance_overall_per_channel)
    )
  })
}

shap_share <- bind_rows(
  read_shap_mode(multi_dir)  |> mutate(mode = "Multiclass"),
  read_shap_mode(binary_dir) |> mutate(mode = "Binary")
) |>
  group_by(mode, config, band) |>
  summarise(value = mean(value), .groups = "drop_last") |>   # mean over seeds
  mutate(share = value / sum(value)) |>                      # normalize per model
  ungroup() |>
  complete(mode, config = config_order, band = band_order) |>  # blanks for unused bands
  mutate(mode   = factor(mode, levels = c("Multiclass", "Binary")),
         config = factor(config, levels = config_order),
         band   = factor(band, levels = rev(band_order)))
```

```{r}
#| label: q5-shap-heatmap
#| fig-width: 16
#| fig-height: 18

shap_share |>
  ggplot(aes(x = config, y = band, fill = share)) +
    geom_tile(colour = "white", linewidth = 1) +
    geom_text(aes(label = percent(share, accuracy = 1), colour = share > 0.10),
              size = 4, na.rm = TRUE, show.legend = FALSE) +
    facet_wrap(~ mode, ncol = 1) +
    scale_colour_manual(values = c(`TRUE` = "white", `FALSE` = "grey20")) +
    seq_fill(labels = percent, name = "Share of model's\ntotal |SHAP|\n(per-channel mean)") +
    scale_x_discrete(labels = config_label) +
    scale_y_discrete(labels = band_label) +
    labs(x = NULL, y = NULL) +
    theme_heatmap
```

Within the Geomorphons band: which landform channels carry the importance
(share of the band's total, seed-mean; forms follow the `rgeomorphon`/GRASS
value order 1-10):

```{r}
#| label: q5-shap-geomorph
#| fig-width: 16
#| fig-height: 12

geomorph_forms <- c("flat", "peak", "ridge", "shoulder", "spur",
                    "slope", "hollow", "footslope", "valley", "pit")

geomorph_shap <- bind_rows(
  read.csv(file.path(multi_dir,  "analysis/shap_geomorph_channels.csv")) |> mutate(mode = "Multiclass"),
  read.csv(file.path(binary_dir, "analysis/shap_geomorph_channels.csv")) |> mutate(mode = "Binary")
) |>
  filter(class == "overall") |>
  group_by(mode, config, geomorphon_form) |>
  summarise(value = mean(mean_abs_shap), .groups = "drop_last") |>
  mutate(share = value / sum(value)) |>
  ungroup() |>
  mutate(mode   = factor(mode, levels = c("Multiclass", "Binary")),
         config = factor(config, levels = config_order),
         geomorphon_form = factor(geomorphon_form, levels = rev(geomorph_forms)))

geomorph_shap |>
  ggplot(aes(x = config, y = geomorphon_form, fill = share)) +
    geom_tile(colour = "white", linewidth = 1) +
    geom_text(aes(label = percent(share, accuracy = 1), colour = share > 0.20),
              size = 4, show.legend = FALSE) +
    facet_wrap(~ mode, ncol = 1) +
    scale_colour_manual(values = c(`TRUE` = "white", `FALSE` = "grey20")) +
    seq_fill(labels = percent, name = "Share of\nGeomorphons\nband |SHAP|") +
    scale_x_discrete(labels = config_label) +
    labs(x = NULL, y = NULL) +
    theme_heatmap
```


### Does the number of patches matter for model accuracy

- F1, IoU, Precision, Recall graphs across the patch numbers input into the model
- Include binary and multiclass stacked graphs
- x-axis is the REALIZED training-set size (from `training_log.json`), points are seeds, lines are seed means
- The `n500` level is dropped: the field-anchored split leaves only ~440-510 training patches, so `n500` realizes essentially the same size as `full` and the two levels overlap.

```{r}
#| label: q6-patchcurve
#| fig-width: 16
#| fig-height: 16

patchcurve <- bind_rows(
  read.csv("../Models/results_patchcurve_v2/multiclass/analysis/patchcurve_long.csv") |> mutate(mode = "Multiclass"),
  read.csv("../Models/results_patchcurve_v2/binary/analysis/patchcurve_long.csv")     |> mutate(mode = "Binary")
) |>
  filter(level != "500") |>   # n500 == full given the ~440-510 patch pool cap
  mutate(mode = factor(mode, levels = c("Multiclass", "Binary")))

pc_class <- patchcurve |>
  pivot_longer(matches("^(iou|f1|precision|recall)_"),
               names_pattern = "^(iou|f1|precision|recall)_(.+)$",
               names_to = c("metric", "class"), values_to = "value") |>
  filter(!is.na(value)) |>   # bind_rows pads the other mode's class columns with NA
  mutate(metric = factor(metric_label[metric], levels = c("Precision", "Recall", "F1", "IoU")),
         class  = factor(class, levels = class_order))

# the realized train count varies a little across seeds within a level,
# so average x and y per level for the mean line
pc_class_mean <- pc_class |>
  group_by(mode, metric, class, level) |>
  summarise(train_patches = mean(train_patches), value = mean(value), .groups = "drop")

pc_class |>
  ggplot(aes(x = train_patches, y = value, colour = class)) +
    geom_point(size = 3, alpha = 0.4) +
    geom_line(data = pc_class_mean, linewidth = 1.2) +
    geom_point(data = pc_class_mean, size = 4) +
    facet_grid(metric ~ mode) +
    scale_colour_manual(values = class_color) +
    labs(x = "Realized training patches", y = NULL) +
    theme_param
```

Model-level curve (macro metrics):

```{r}
#| label: q6-patchcurve-macro
#| fig-width: 16
#| fig-height: 8

pc_macro <- patchcurve |>
  pivot_longer(c(macro_f1, mean_iou, overall_accuracy), names_to = "metric", values_to = "value") |>
  mutate(metric = factor(metric_label[metric], levels = c("Macro F1", "Mean IoU", "Overall accuracy")))

pc_macro_mean <- pc_macro |>
  group_by(mode, metric, level) |>
  summarise(train_patches = mean(train_patches), value = mean(value), .groups = "drop")

pc_macro |>
  ggplot(aes(x = train_patches, y = value, colour = metric)) +
    geom_point(size = 3, alpha = 0.4) +
    geom_line(data = pc_macro_mean, linewidth = 1.2) +
    geom_point(data = pc_macro_mean, size = 4) +
    facet_grid(. ~ mode) +
    scale_colour_manual(values = c("Macro F1" = "#388C73", "Mean IoU" = "#6a4836",
                                   "Overall accuracy" = "grey40"), name = NULL) +
    labs(x = "Realized training patches", y = NULL) +
    theme_param
```


### Does model architecture matter?

- Boxplot of metrics with point overlaid to show distribution of model runs
- U-Net vs UNet3+ on `fld_chmret_leafoff` (same seeds -> same test patches; bf64/d5 held constant)

```{r}
#| label: q7-arch
#| fig-width: 16
#| fig-height: 12

arch_color <- c("U-Net" = "#4477AA", "UNet3+" = "#EE7733")

arch_data <- bind_rows(
  read.csv("../Models/results_arch_v2/multiclass/analysis/arch_compare.csv") |> mutate(mode = "Multiclass"),
  read.csv("../Models/results_arch_v2/binary/analysis/arch_compare.csv")     |> mutate(mode = "Binary")
) |>
  filter(seed != "mean") |>
  select(-starts_with("arch_"), -starts_with("delta_")) |>
  pivot_longer(-c(config, seed, mode),
               names_pattern = "^(.*)_(unet3plus|unet)$",
               names_to = c("metric", "arch"), values_to = "value") |>
  filter(!is.na(value)) |>   # bind_rows pads the other mode's class columns with NA
  mutate(mode = factor(mode, levels = c("Multiclass", "Binary")),
         arch = factor(if_else(arch == "unet", "U-Net", "UNet3+"), levels = names(arch_color)))

arch_macro <- arch_data |>
  filter(metric %in% names(metric_label)) |>
  mutate(metric = factor(metric_label[metric], levels = c("Macro F1", "Mean IoU", "Overall accuracy"))) |>
  ggplot(aes(x = metric, y = value, fill = arch)) +
    box_jitter(point_size = 4) +
    facet_grid(. ~ mode) +
    scale_fill_manual(values = arch_color, name = NULL) +
    labs(x = NULL, y = NULL, title = "Model-level metrics") +
    theme_param

arch_class <- arch_data |>
  filter(str_starts(metric, "iou_")) |>
  mutate(class = factor(str_remove(metric, "^iou_"), levels = class_order)) |>
  ggplot(aes(x = class, y = value, fill = arch)) +
    box_jitter(point_size = 4) +
    facet_grid(. ~ mode, scales = "free_x", space = "free_x") +
    scale_fill_manual(values = arch_color, name = NULL) +
    labs(x = NULL, y = "IoU", title = "Per-class IoU") +
    theme_param

arch_macro / arch_class + plot_layout(guides = "collect")
```
