---
title: "Factorial Experiment Visualization (v3)"
author: "Anthony Stewart"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
    fig-width: 16
    fig-height: 9
execute:
  warning: false
  message: false
---

Walkthrough of the v3 factorial experiment results (8 configs x 5 seeds x
{multiclass, binary}, 29-channel full feature set) plus the three-arm
architecture comparison (U-Net / UNet3+ / mbfusion) and the fusion gate exports.
Inputs are the small CSV/JSON aggregation outputs, plus the float16 gate rasters
for the last section:

- `dl_08_aggregate_factorial.py` -> `Models/factorial_results_v3/<mode>/analysis/`
- `dl_09_shap_factorial.py` -> `Models/factorial_results_v3/<mode>/<config>/seed<k>/shap/*_shap_importance.json`
- `dl_09b_shap_geomorph_export.py` -> `Models/factorial_results_v3/<mode>/analysis/shap_geomorph_channels.csv`
- `dl_08b_aggregate_patchcurve.py --arch-compare` -> `Models/results_arch_fusion_v3/<mode>/analysis/`
  (it writes into the LAST `--arch-dir`, which is the fusion root)
- `dl_11_export_gates.py` -> `Models/results_arch_fusion_v3/<mode>/*_mbfusion/seed<k>/gates/`

The patch-count learning curve was dropped from v3 (the 100-500 range is under one
order of magnitude and will not support a scaling claim), so there is no Q6 curve
here; the v2 curve reproduces from the frozen `results_patchcurve_v2` tree.

```{r}
#| label: setup

library(tidyverse)
library(patchwork)
library(jsonlite)
library(colorspace)
library(scales)
library(terra)     # only for the fusion gate maps: reads MOD_CLASS off the patches
```


```{r}
#| label: factorial-data

multi_dir  <- "../Models/factorial_results_v3/multiclass"
binary_dir <- "../Models/factorial_results_v3/binary"

multi_fact_data <- read.csv(file.path(multi_dir, "analysis/factorial_long.csv"))
glimpse(multi_fact_data)

binary_fact_data <- read.csv(file.path(binary_dir, "analysis/factorial_long.csv"))
glimpse(binary_fact_data)
```

```{r}
#| label: results-roots

# Root holding the architecture-comparison analysis/ dirs -- i.e. the LAST
# --arch-dir passed to dl_08b_aggregate_patchcurve.py, which is where it writes.
# For v3 that is the fusion root, which also holds the gate exports (the .npz +
# gate_summary.json under <cell>/seed<k>/gates/) read by the final section.
arch_dir_base <- "../Models/results_arch_fusion_v3"

# Field patches, for pairing gate rasters back to their MOD_CLASS labels.
patch_dir <- "../Data/Training_Data/R_Patches"
```

```{r}
#| label: data-organizing

config_order <- c("fld_nolidar_leafon", "fld_nolidar_leafoff",  "fld_chmret_leafon", "fld_chmret_leafoff",
                 "nwi_chmret_leafoff", "nwiextra_chmret_leafoff", "nwifield_chmret_leafoff", "flddeg_chmret_leafoff")

config_label <- c("fld_nolidar_leafon" = "Baseline",
                  "fld_nolidar_leafoff" = "With\nLeaf-Off",
                  "fld_chmret_leafon" = "With\nLidar",
                  "fld_chmret_leafoff" = "Full\nModel",
                  "nwi_chmret_leafoff" = "NWI\nData",
                  "nwiextra_chmret_leafoff" = "NWI\nx2",
                  "nwifield_chmret_leafoff" = "NWI\n+\nField",
                  "flddeg_chmret_leafoff" = "Degraded\nField")

# single-line variant for heatmap y-axes (stacked labels overlap there)
config_label_1line <- gsub("\n ", ", ", config_label)

class_order <- c("EMW", "FSW", "SSW", "WET", "UPL")
class_color <- c("FSW" = "#81b14f", "SSW" = "#c4bd8b", "EMW" = "#8faab3", "UPL" = "#6a4836", "WET" = "#388C73")

# Which multiclass classes collapse into WET, read from the pipeline's own config
# so the notebook cannot drift from what the binary models were trained on.
binary_wet <- fromJSON("../Python_Code_Analysis/DL_Pipeline_v2/dl_band_config.json"
                       )$binary_mapping$WET

metric_label <- c("precision" = "Precision", "recall" = "Recall", "f1" = "F1", "iou" = "IoU",
                  "macro_f1" = "Macro F1", "mean_iou" = "Mean IoU", "overall_accuracy" = "Overall accuracy")

organize <- function(df) {
  df |>
    mutate(config = factor(config, levels = config_order),
           class  = factor(class,  levels = class_order))
}
multi_fact_data  <- organize(multi_fact_data)
binary_fact_data <- organize(binary_fact_data)

# Both modes in one frame for faceted figures
fact_data <- bind_rows(
  multi_fact_data  |> mutate(mode = "Multiclass"),
  binary_fact_data |> mutate(mode = "Binary")
) |> mutate(mode = factor(mode, levels = c("Multiclass", "Binary")))

theme_param <- theme(
  text = element_text(size = 18),
  legend.position = 'right',
  legend.text = element_text(size = rel(0.70), hjust = 0.5),
  panel.background = element_blank(),
  panel.grid.major = element_line(colour = "grey50", linewidth = 0.2),
  axis.ticks = element_blank(),
  axis.text = element_text(size = rel(0.9)),
  axis.title = element_text(size = rel(1)),
  strip.background = element_blank(),
  strip.text = element_text(size = rel(0.9), face = "bold")
)

theme_heatmap <- theme_param + theme(panel.grid.major = element_blank())

# Boxplot + jittered seed points, shared across the metric figures
box_jitter <- function(point_size = 5) {
  list(
    geom_boxplot(outlier.shape = NA, position = position_dodge(width = 0.75)),
    geom_jitter(position = position_jitterdodge(jitter.width = 0.15, dodge.width = 0.75),
                size = point_size, shape = 21, colour = "grey15")
  )
}

# Single-hue sequential fill for the heatmaps (magnitude = light -> dark)
seq_fill <- function(...) {
  scale_fill_gradient(low = "#f2f7f5", high = "#14523f", na.value = "grey93", ...)
}
```

### Does leaf off imagery or lidar-derived metrics improve field-based models?

Structure

- Just using the the field-verified dataset models
- Model config on the x-axis, model evaluation metric on the y-axis
- box and whiskers plots representing the model measure spread, points for the individual model measures
- Precision and recall share a panel: full class color = precision, lighter shade = recall

```{r}
#| label: q1-precision-recall
#| fig-width: 8
#| fig-height: 14
#| layout-ncol: 2

pr_data <- fact_data |>
  filter(label == "fld") |>
  pivot_longer(c(precision, recall), names_to = "metric", values_to = "value") |>
  mutate(metric = factor(metric_label[metric], levels = c("Precision", "Recall")))

pr_plot <- function(which_mode, which_metric) {
  pr_data |>
    filter(mode == which_mode, metric == which_metric) |>
    ggplot(aes(x = config, y = value, fill = class)) +
      facet_wrap(~ class, scales = "free_y", ncol = 1) +
      box_jitter(point_size = 3.5) +
      scale_fill_manual(values = class_color) +
      scale_x_discrete(labels = config_label) +
      labs(x = NULL, y = which_metric, title = which_mode) +
      theme_param
}

pr_plot("Multiclass", "Precision")

pr_plot("Multiclass", "Recall")

pr_plot("Binary", "Precision")

pr_plot("Binary", "Recall")
```

```{r}
#| label: q1-f1-iou
#| fig-width: 8
#| fig-height: 14

fact_data |>
  filter(label == "fld", mode == "Multiclass") |>
  ggplot(aes(x = config, y = f1, fill = class)) +
    box_jitter() +
    facet_wrap(~ class, scales = "free_y", ncol = 1) +
    scale_fill_manual(values = class_color) +
    scale_x_discrete(labels = config_label) +
    scale_y_continuous(labels = scales::label_number(accuracy = 0.01)) +
    labs(x = NULL, y = "F1") +
    # ylim(0.0, 1.0) +
    theme_param

fact_data |>
  filter(label == "fld", mode == "Multiclass") |>
  ggplot(aes(x = config, y = iou, fill = class)) +
    box_jitter() +
    facet_wrap(~ class, scales = "free_y", ncol = 1) +
    scale_fill_manual(values = class_color) +
    scale_x_discrete(labels = config_label) +
    scale_y_continuous(labels = scales::label_number(accuracy = 0.01)) +
    labs(x = NULL, y = "IoU") +
    theme_param

  fact_data |>
  filter(label == "fld", mode == "Binary") |>
  ggplot(aes(x = config, y = f1, fill = class)) +
    box_jitter() +
    facet_wrap(~ class, scales = "free_y", ncol = 1) +
    scale_fill_manual(values = class_color) +
    scale_x_discrete(labels = config_label) +
    scale_y_continuous(labels = scales::label_number(accuracy = 0.01)) +
    labs(x = NULL, y = "F1") +
    theme_param

fact_data |>
  filter(label == "fld", mode == "Binary") |>
  ggplot(aes(x = config, y = iou, fill = class)) +
    box_jitter() +
    facet_wrap(~ class, scales = "free_y", ncol = 1) +
    scale_fill_manual(values = class_color) +
    scale_x_discrete(labels = config_label) +
    scale_y_continuous(labels = scales::label_number(accuracy = 0.01)) +
    labs(x = NULL, y = "IoU") +
    theme_param
```


### How do NWI data labels compare to field and GIS-derived data?

#### Also does degrading the field data hinder its performance?

- All label sources on the full feature set (`chmret_leafoff`), evaluated against field labels

```{r}
#| label: q2-nwi-f1-iou
#| fig-width: 24
#| fig-height: 6

config_label_nwi <- config_label
config_label_nwi[4] <- "Full\nModel\nField+GIS"

fact_data |>
  filter(featureset == "chmret_leafoff", mode == "Multiclass") |>
  ggplot(aes(x = config, y = f1, fill = class)) +
    box_jitter() +
    facet_wrap(mode ~ class, scales = "free_y", ncol = 4) +
    scale_fill_manual(values = class_color) +
    scale_x_discrete(labels = config_label_nwi) +
    scale_y_continuous(labels = scales::label_number(accuracy = 0.01)) +
    labs(x = NULL, y = "F1") +
    theme_param

fact_data |>
  filter(featureset == "chmret_leafoff", mode == "Binary") |>
  ggplot(aes(x = config, y = f1, fill = class)) +
    box_jitter() +
    facet_wrap(mode ~ class, scales = "free_y", ncol = 4) +
    scale_fill_manual(values = class_color) +
    scale_x_discrete(labels = config_label) +
    scale_y_continuous(labels = scales::label_number(accuracy = 0.01)) +
    labs(x = NULL, y = "F1") +
    theme_param

# fact_data |>
#   filter(featureset == "chmret_leafoff", mode == "Multiclass") |>
#   ggplot(aes(x = config, y = iou, fill = class)) +
#     box_jitter() +
#     facet_wrap(mode ~ class, scales = "free_y", ncol = 4) +
#     scale_fill_manual(values = class_color) +
#     scale_x_discrete(labels = config_label) +
#     scale_y_continuous(labels = scales::label_number(accuracy = 0.01)) +
#     labs(x = NULL, y = "IoU") +
#     theme_param

# fact_data |>
#   filter(featureset == "chmret_leafoff", mode == "Binary") |>
#   ggplot(aes(x = config, y = iou, fill = class)) +
#     box_jitter() +
#     facet_wrap(mode ~ class, scales = "free_y", ncol = 4) +
#     scale_fill_manual(values = class_color) +
#     scale_x_discrete(labels = config_label) +
#     scale_y_continuous(labels = scales::label_number(accuracy = 0.01)) +
#     labs(x = NULL, y = "IoU") +
#     theme_param
```

Model-level summary (macro metrics are repeated on every class row of
`factorial_long.csv`, so collapse to one row per model first):

```{r}
#| label: q2-macro
#| fig-width: 20
#| fig-height: 8

fact_data |>
  filter(featureset == "chmret_leafoff", mode == "Multiclass") |>
  distinct(mode, config, seed, macro_f1) |>
  pivot_longer(c(macro_f1), names_to = "metric", values_to = "value") |>
  mutate(metric = metric_label[metric]) |>
  ggplot(aes(x = config, y = value, fill = metric)) +
    box_jitter(point_size = 4) +
    facet_grid(. ~ mode) +
    scale_fill_manual(values = c("Macro F1" = "#388C73", "Mean IoU" = lighten("#388C73", 0.45)), name = NULL) +
    scale_x_discrete(labels = config_label) +
    labs(x = NULL, y = NULL) +
    theme_param
  
fact_data |>
  filter(featureset == "chmret_leafoff", mode == "Multiclass") |>
  distinct(mode, config, seed, mean_iou) |>
  pivot_longer(c(mean_iou), names_to = "metric", values_to = "value") |>
  mutate(metric = metric_label[metric]) |>
  ggplot(aes(x = config, y = value, fill = metric)) +
    box_jitter(point_size = 4) +
    facet_grid(. ~ mode) +
    scale_fill_manual(values = c("Macro F1" = "#388C73", "Mean IoU" = lighten("#388C73", 0.45)), name = NULL) +
    scale_x_discrete(labels = config_label) +
    labs(x = NULL, y = NULL) +
    theme_param

fact_data |>
  filter(featureset == "chmret_leafoff", mode == "Binary") |>
  distinct(mode, config, seed, macro_f1) |>
  pivot_longer(c(macro_f1), names_to = "metric", values_to = "value") |>
  mutate(metric = metric_label[metric]) |>
  ggplot(aes(x = config, y = value, fill = metric)) +
    box_jitter(point_size = 4) +
    facet_grid(. ~ mode) +
    scale_fill_manual(values = c("Macro F1" = "#388C73", "Mean IoU" = lighten("#388C73", 0.45)), name = NULL) +
    scale_x_discrete(labels = config_label) +
    labs(x = NULL, y = NULL) +
    theme_param
  
fact_data |>
  filter(featureset == "chmret_leafoff", mode == "Binary") |>
  distinct(mode, config, seed, mean_iou) |>
  pivot_longer(c(mean_iou), names_to = "metric", values_to = "value") |>
  mutate(metric = metric_label[metric]) |>
  ggplot(aes(x = config, y = value, fill = metric)) +
    box_jitter(point_size = 4) +
    facet_grid(. ~ mode) +
    scale_fill_manual(values = c("Macro F1" = "#388C73", "Mean IoU" = lighten("#388C73", 0.45)), name = NULL) +
    scale_x_discrete(labels = config_label) +
    labs(x = NULL, y = NULL) +
    theme_param
```


### How do field and NWI models compare across each other

- Heat maps of model metrics: config (rows) x class (cols), one panel per metric.
- First the absolute seed-mean scores, then the same grid as a **difference from a
  reference config** -- e.g. multiclass EMW precision for `With Leaf-Off` = 0.54
  minus 0.49 for `Baseline` = +0.05∆.

```{r}
#| label: q3-metric-heatmap
#| fig-width: 18
#| fig-height: 14

metric_heat <- fact_data |>
  pivot_longer(c(precision, recall, f1, iou), names_to = "metric", values_to = "value") |>
  mutate(metric = factor(metric_label[metric], levels = c("Precision", "Recall", "F1", "IoU"))) |>
  group_by(mode, config, class, metric) |>
  summarise(value = mean(value), .groups = "drop")

heat_panel <- function(which_mode) {
  metric_heat |>
    filter(mode == which_mode) |>
    ggplot(aes(x = class, y = fct_rev(config), fill = value)) +
      geom_tile(colour = "white", linewidth = 1) +
      geom_text(aes(label = number(value, accuracy = 0.01),
                    colour = value > 0.55), size = 4.5, show.legend = FALSE) +
      facet_grid(. ~ metric) +
      scale_colour_manual(values = c(`TRUE` = "white", `FALSE` = "grey20")) +
      seq_fill(limits = c(0, 1), name = "Seed-mean\nscore") +
      scale_y_discrete(labels = config_label_1line) +
      labs(x = NULL, y = NULL, title = which_mode) +
      theme_heatmap
}

heat_panel("Multiclass") / heat_panel("Binary") +
  plot_layout(guides = "collect", heights = c(1, 1))
```

The same grid as a difference from a reference config. Deltas are computed
**per seed** and then averaged -- same seed => same test patches, so each seed
contributes a genuinely paired difference -- and a cell is outlined when all
`r n_distinct(fact_data$seed)` seeds move the same way, which at n=5 is the
credible summary rather than a p-value. Each panel gets its own colour scale --
the binary deltas are roughly an order of magnitude smaller than the multiclass
ones, so a shared scale would render the binary panel blank.

Two references are worth reading, so `HEAT_REF` switches between them:
`fld_nolidar_leafon` (the "Baseline" model) reads the **feature** rows as "what
did lidar and leaf-off buy", but the four NWI rows below it change labels *and*
features at once, so those are a compound difference. `fld_chmret_leafoff` (the
"Full Model") isolates the **label gradient**, since every config it is compared
against there shares the identical 29-channel input set.

```{r}
#| label: q3-metric-heatmap-delta
#| fig-width: 18
#| fig-height: 14

HEAT_REF <- "fld_nolidar_leafon"   # or "fld_chmret_leafoff" for the label gradient

metric_delta <- fact_data |>
  pivot_longer(c(precision, recall, f1, iou), names_to = "metric", values_to = "value") |>
  mutate(metric = factor(metric_label[metric], levels = c("Precision", "Recall", "F1", "IoU"))) |>
  # paired within seed: join each config's score to the reference's same-seed score
  left_join(
    fact_data |>
      filter(config == HEAT_REF) |>
      pivot_longer(c(precision, recall, f1, iou), names_to = "metric", values_to = "ref") |>
      mutate(metric = factor(metric_label[metric], levels = c("Precision", "Recall", "F1", "IoU"))) |>
      select(mode, seed, class, metric, ref),
    by = c("mode", "seed", "class", "metric")) |>
  filter(config != HEAT_REF, !is.na(ref)) |>
  mutate(delta = value - ref) |>
  group_by(mode, config, class, metric) |>
  # sign_consistent BEFORE delta is overwritten -- summarise() evaluates in order,
  # so testing it afterwards would just re-test the mean and mark every cell.
  summarise(sign_consistent = all(delta > 0) | all(delta < 0),
            delta = mean(delta),
            .groups = "drop")

heat_delta_panel <- function(which_mode) {
  d <- filter(metric_delta, mode == which_mode)
  # Limits per panel: the binary deltas are an order of magnitude smaller than the
  # multiclass ones, so a shared scale would render that panel uniformly blank.
  lim <- max(abs(d$delta), na.rm = TRUE)
  ggplot(d, aes(x = class, y = fct_rev(config), fill = delta)) +
    geom_tile(colour = "white", linewidth = 1) +
    # outline = all seeds moved the same way
    geom_tile(data = filter(d, sign_consistent), fill = NA, colour = "grey15", linewidth = 1.1) +
    geom_text(aes(label = sprintf("%+.2f", delta), colour = abs(delta) > lim * 0.55),
              size = 4.5, show.legend = FALSE) +
    facet_grid(. ~ metric) +
    scale_colour_manual(values = c(`TRUE` = "white", `FALSE` = "grey20")) +
    scale_fill_gradient2(low = "#8c510a", mid = "grey96", high = "#14523f", midpoint = 0,
                         limits = c(-lim, lim),
                         name = paste0("\u0394 vs ", config_label_1line[[HEAT_REF]],
                                       "\n(", which_mode, ")")) +
    scale_y_discrete(labels = config_label_1line) +
    labs(x = NULL, y = NULL, title = which_mode) +
    theme_heatmap
}

# Deliberately NOT guides = "collect": each panel carries its own delta scale.
heat_delta_panel("Multiclass") / heat_delta_panel("Binary") +
  plot_layout(heights = c(1, 1))
```


### How accurate are the classes via confusion matrices?

- binary and multiclass model confusion matrices for each configuration
- color and fill by relative % but include a parentheses with the number of pixels e.g. 70% (3,400,000)
- rows (true class) sum to 100%, so the diagonal reads as recall; counts are seed-mean pixels

```{r}
#| label: q4-confusion-load

read_confusion <- function(mode_dir) {
  map_dfr(config_order, function(cfg) {
    f <- file.path(mode_dir, "analysis/confusion_mean", paste0(cfg, ".csv"))
    if (!file.exists(f)) return(NULL)
    read.csv(f, row.names = 1, check.names = FALSE) |>
      rownames_to_column("true") |>
      pivot_longer(-true, names_to = "pred", values_to = "n") |>
      mutate(config = factor(cfg, levels = config_order))
  }) |>
    group_by(config, true) |>
    mutate(prop = n / sum(n)) |>
    ungroup() |>
    mutate(true = factor(true, levels = rev(class_order)),
           pred = factor(pred, levels = class_order),
           cell_label = paste0(percent(prop, accuracy = 1), "\n(",
                               number(n, accuracy = 0.1, scale_cut = cut_short_scale()), ")"))
}

confusion_plot <- function(mode_dir, title, text_size = 4.5) {
  read_confusion(mode_dir) |>
    ggplot(aes(x = pred, y = true, fill = prop)) +
      geom_tile(colour = "white", linewidth = 1) +
      geom_text(aes(label = cell_label, colour = prop > 0.55),
                size = text_size, lineheight = 0.9, show.legend = FALSE) +
      facet_wrap(~ config, ncol = 4, labeller = labeller(config = config_label)) +
      scale_colour_manual(values = c(`TRUE` = "white", `FALSE` = "grey20")) +
      seq_fill(limits = c(0, 1), labels = percent, name = "% of\ntrue class") +
      labs(x = "Predicted", y = "True", title = title) +
      theme_heatmap
}
```

```{r}
#| label: q4-confusion-multi
#| fig-width: 20
#| fig-height: 11

confusion_plot(multi_dir, "Multiclass")
```

```{r}
#| label: q4-confusion-binary
#| fig-width: 20
#| fig-height: 8

confusion_plot(binary_dir, "Binary", text_size = 5.5)
```


### What were the important features in each model configuration according to SHAP?

- Model heatmap with cross-model comparison: each model normalized to its own total importance, so bands are comparable across configs. Bands a config doesn't use are blank.
- Uses the PER-CHANNEL MEAN aggregation (not the raw sum) so the one-hot Geomorph_local band is not inflated 10x relative to single-channel bands.
- Also show within the Geomorphons category a heatmap of the different categories across models

```{r}
#| label: q5-shap-load

# The v3 predictor set: 20 bands -> 29 channels. TPI_local / meanc_local /
# dmv_local are v3 additions (v2 had 17 bands), so this list is longer than the
# v2 notebook's. Order = terrain, geomorphon, lidar structure, leaf-on, leaf-off.
band_order <- c("DEM", "slope_local", "TPI_local", "meanc_local", "dmv_local",
                "Geomorph_local", "flowacc", "twi",
                "CHM", "pct_below_1m", "pct_1m_to_5m", "pct_above_5m",
                "r", "g", "b", "nir", "r_lo", "g_lo", "b_lo", "nir_lo")

band_label <- c("DEM" = "DEM", "slope_local" = "Slope", "TPI_local" = "TPI",
                "meanc_local" = "Mean curvature", "dmv_local" = "Deviation from\nmean elevation",
                "Geomorph_local" = "Geomorphons",
                "flowacc" = "Flow accum.", "twi" = "TWI",
                "CHM" = "CHM", "pct_below_1m" = "% returns < 1 m",
                "pct_1m_to_5m" = "% returns 1-5 m", "pct_above_5m" = "% returns > 5 m",
                "r" = "Red (leaf-on)", "g" = "Green (leaf-on)", "b" = "Blue (leaf-on)", "nir" = "NIR (leaf-on)",
                "r_lo" = "Red (leaf-off)", "g_lo" = "Green (leaf-off)", "b_lo" = "Blue (leaf-off)", "nir_lo" = "NIR (leaf-off)")

read_shap_mode <- function(mode_dir) {
  files <- Sys.glob(file.path(mode_dir, "*", "seed*", "shap", "*_shap_importance.json"))
  map_dfr(files, function(f) {
    j <- fromJSON(f)
    tibble(
      config = basename(dirname(dirname(dirname(f)))),
      seed   = parse_number(str_extract(f, "seed\\d+")),
      band   = names(j$importance_overall_per_channel),
      value  = unlist(j$importance_overall_per_channel)
    )
  })
}

shap_share <- bind_rows(
  read_shap_mode(multi_dir)  |> mutate(mode = "Multiclass"),
  read_shap_mode(binary_dir) |> mutate(mode = "Binary")
) |>
  group_by(mode, config, band) |>
  summarise(value = mean(value), .groups = "drop_last") |>   # mean over seeds
  mutate(share = value / sum(value)) |>                      # normalize per model
  ungroup() |>
  complete(mode, config = config_order, band = band_order) |>  # blanks for unused bands
  mutate(mode   = factor(mode, levels = c("Multiclass", "Binary")),
         config = factor(config, levels = config_order),
         band   = factor(band, levels = rev(band_order)))
```

```{r}
#| label: q5-shap-heatmap
#| fig-width: 16
#| fig-height: 18

shap_share |>
  ggplot(aes(x = config, y = band, fill = share)) +
    geom_tile(colour = "white", linewidth = 1) +
    geom_text(aes(label = percent(share, accuracy = 1), colour = share > 0.10),
              size = 4, na.rm = TRUE, show.legend = FALSE) +
    facet_wrap(~ mode, ncol = 1) +
    scale_colour_manual(values = c(`TRUE` = "white", `FALSE` = "grey20")) +
    seq_fill(labels = percent, name = "Share of model's\ntotal |SHAP|\n(per-channel mean)") +
    scale_x_discrete(labels = config_label) +
    scale_y_discrete(labels = band_label) +
    labs(x = NULL, y = NULL) +
    theme_heatmap
```

Within the Geomorphons band: which landform channels carry the importance
(share of the band's total, seed-mean; forms follow the `rgeomorphon`/GRASS
value order 1-10):

```{r}
#| label: q5-shap-geomorph
#| fig-width: 16
#| fig-height: 12

geomorph_forms <- c("flat", "peak", "ridge", "shoulder", "spur",
                    "slope", "hollow", "footslope", "valley", "pit")

# dl_09b_shap_geomorph_export.py has to be re-run per results root; skip the
# panel rather than fail the render if v3's CSVs are not there yet.
geomorph_csv <- file.path(c(multi_dir, binary_dir), "analysis/shap_geomorph_channels.csv")
have_geomorph <- all(file.exists(geomorph_csv))
if (!have_geomorph) {
  cat("shap_geomorph_channels.csv not found under\n  ",
      paste(geomorph_csv, collapse = "\n   "),
      "\nRun: python Python_Code_Analysis/DL_Pipeline_v2/dl_09b_shap_geomorph_export.py",
      "--results-dir Models/factorial_results_v3/<mode>\n")
}

geomorph_shap <- if (!have_geomorph) NULL else bind_rows(
  read.csv(file.path(multi_dir,  "analysis/shap_geomorph_channels.csv")) |> mutate(mode = "Multiclass"),
  read.csv(file.path(binary_dir, "analysis/shap_geomorph_channels.csv")) |> mutate(mode = "Binary")
) |>
  filter(class == "overall") |>
  group_by(mode, config, geomorphon_form) |>
  summarise(value = mean(mean_abs_shap), .groups = "drop_last") |>
  mutate(share = value / sum(value)) |>
  ungroup() |>
  mutate(mode   = factor(mode, levels = c("Multiclass", "Binary")),
         config = factor(config, levels = config_order),
         geomorphon_form = factor(geomorphon_form, levels = rev(geomorph_forms)))

if (have_geomorph) geomorph_shap |>
  ggplot(aes(x = config, y = geomorphon_form, fill = share)) +
    geom_tile(colour = "white", linewidth = 1) +
    geom_text(aes(label = percent(share, accuracy = 1), colour = share > 0.20),
              size = 4, show.legend = FALSE) +
    facet_wrap(~ mode, ncol = 1) +
    scale_colour_manual(values = c(`TRUE` = "white", `FALSE` = "grey20")) +
    seq_fill(labels = percent, name = "Share of\nGeomorphons\nband |SHAP|") +
    scale_x_discrete(labels = config_label) +
    labs(x = NULL, y = NULL) +
    theme_heatmap
```


### Does model architecture matter?

Three arms, all on `fld_chmret_leafoff` (the 29-channel full feature set), both
modes, the same 5 seeds, and bf64/d5/50-epochs held constant -- so the only thing
that varies is the network. No arm is pre-trained: the v2 arms were 26-channel
runs on the old pool and are not comparable, so all three were retrained for v3.

| Arm | What it changes | Results root |
|---|---|---|
| **U-Net** | the reference: residual blocks + SE attention | `Models/factorial_results_v3/<mode>/fld_chmret_leafoff/` |
| **UNet3+** | full-scale skips: every decoder node sees every encoder scale | `Models/results_arch_v3/<mode>/<config>_unet3plus/` |
| **mbfusion** | one encoder per input modality, fused by a per-pixel softmax gate at all 6 scales, into a decoder bit-identical to the U-Net's | `Models/results_arch_fusion_v3/<mode>/<config>_mbfusion/` |

Figure layout, per mode: model-level Macro F1 and Mean IoU on top; per-class F1
coloured by class in the middle; the FSW/UPL confusion rates the fusion encoder
is specifically meant to attack on the bottom (multiclass only -- binary has no
FSW, so `dl_08b` omits those columns rather than writing zeros). Boxplot with the
five seed points overlaid, since at n=5 the spread is the story.

Everything reads `arch_compare_long.csv` (tidy, one row per arch x seed), so a
fourth arm added to `dl_08b_aggregate_patchcurve.py --arch-dir` needs nothing here
beyond a display name and a colour. Arms absent from the CSV drop out silently.

```{r}
#| label: q7-arch-data

# Display names + palette, keyed by the `arch` values dl_08b writes. Extend both
# to add an arm; everything downstream keys off ARCH_LEVELS.
arch_name  <- c(unet = "U-Net", unet3plus = "UNet3+", mbfusion = "Fusion")
arch_color <- c("U-Net" = "#4477AA", "UNet3+" = "#EE7733", "Fusion" = "#009988")

arch_root <- function(mode_dir) file.path(arch_dir_base, mode_dir, "analysis")

read_arch_csv <- function(file) {
  bind_rows(
    read.csv(file.path(arch_root("multiclass"), file)) |> mutate(mode = "Multiclass"),
    read.csv(file.path(arch_root("binary"), file))     |> mutate(mode = "Binary")
  ) |>
    mutate(mode = factor(mode, levels = c("Multiclass", "Binary")))
}

arch_long <- read_arch_csv("arch_compare_long.csv") |>
  mutate(arch = factor(arch_name[arch], levels = arch_name))

# Levels actually present, so a two-arm (pre-fusion) tree still renders
ARCH_LEVELS <- levels(droplevels(arch_long$arch))
arch_long <- arch_long |> mutate(arch = factor(as.character(arch), levels = ARCH_LEVELS))

# Long-form metrics: macro/mean-IoU, per-class IoU & F1, and confusion rates all
# arrive as columns; melt everything numeric and label the pieces.
arch_metrics <- arch_long |>
  select(-any_of(c("params", "gflops", "sec_per_epoch", "epochs_run", "fit_seconds"))) |>
  pivot_longer(where(is.numeric) & !any_of("seed"),
               names_to = "metric", values_to = "value") |>
  filter(!is.na(value))

arch_class <- arch_metrics |>
  filter(str_detect(metric, "^(iou|f1)_")) |>
  separate_wider_regex(metric, c(metric = "iou|f1", "_", class = ".*")) |>
  mutate(class = factor(class, levels = class_order))
```

```{r}
#| label: q7-arch
#| fig-width: 16
#| fig-height: 16

# Alpha separates arms inside a class group; recycled to however many arms exist.
arch_alpha <- setNames(seq(1, 0.4, length.out = length(ARCH_LEVELS)), ARCH_LEVELS)

arch_fig <- function(m) {
  # Macro F1 and Mean IoU side by side: F1 is the headline, IoU is the stricter
  # read and the one the factorial's own summary tables lead with.
  macro <- arch_metrics |>
    filter(metric %in% c("macro_f1", "mean_iou"), mode == m) |>
    mutate(metric = factor(metric_label[metric], levels = c("Macro F1", "Mean IoU"))) |>
    ggplot(aes(x = arch, y = value, fill = arch)) +
      box_jitter(point_size = 4) +
      facet_wrap(~ metric, scales = "free_y") +
      scale_fill_manual(values = arch_color, name = NULL) +
      labs(x = NULL, y = NULL, title = paste0(m, ": model-level accuracy")) +
      theme_param

  cls <- arch_class |>
    filter(metric == "f1", mode == m) |>
    ggplot(aes(x = class, y = value, fill = class, alpha = arch,
               group = interaction(class, arch))) +
      box_jitter(point_size = 4) +
      scale_fill_manual(values = class_color, guide = "none") +
      scale_alpha_manual(values = arch_alpha, name = NULL) +
      guides(alpha = guide_legend(override.aes = list(fill = "grey30"))) +
      labs(x = NULL, y = "F1",
           title = paste0("Per-class F1 (", paste(ARCH_LEVELS, collapse = " vs "), ")")) +
      theme_param

  parts <- list(macro, cls)
  heights <- c(1, 1.4)

  # Confusion panel: multiclass only -- binary has no FSW, so dl_08b omits the
  # conf_ columns there rather than writing zeros. `_swap` is the symmetric total
  # (both directions over both classes' pixels), so it moves less than either leg.
  conf_label <- c(conf_FSW_as_UPL   = "True FSW -> predicted UPL",
                  conf_UPL_as_FSW   = "True UPL -> predicted FSW",
                  conf_FSW_UPL_swap = "FSW <-> UPL swap (both directions)")
  conf <- arch_metrics |> filter(mode == m, metric %in% names(conf_label))
  if (nrow(conf) > 0) {
    conf_p <- conf |>
      mutate(metric = factor(conf_label[metric], levels = conf_label[names(conf_label)])) |>
      ggplot(aes(x = arch, y = value, fill = arch)) +
        box_jitter(point_size = 4) +
        facet_wrap(~ metric, scales = "free_y") +
        scale_fill_manual(values = arch_color, guide = "none") +
        scale_y_continuous(labels = scales::label_percent(accuracy = 0.1)) +
        labs(x = NULL, y = "Share of the true class's pixels",
             title = "FSW <-> UPL confusion (lower is better)") +
        theme_param
    parts <- c(parts, list(conf_p))
    heights <- c(heights, 1)
  }

  wrap_plots(parts, ncol = 1) + plot_layout(heights = heights)
}

arch_fig("Multiclass")
arch_fig("Binary")
```

Seed-mean scores per arm, the table behind the boxplots:

```{r}
#| label: q7-arch-summary

arch_metrics |>
  filter(metric %in% c("macro_f1", "mean_iou", "overall_accuracy") |
         str_detect(metric, "^(f1|iou)_") | str_detect(metric, "^conf_")) |>
  group_by(mode, metric, arch) |>
  summarise(mean = mean(value), sd = sd(value), .groups = "drop") |>
  mutate(cell = sprintf("%.3f (%.3f)", mean, sd)) |>
  select(mode, metric, arch, cell) |>
  pivot_wider(names_from = arch, values_from = cell) |>
  arrange(mode, metric) |>
  knitr::kable(caption = "Seed mean (SD) per architecture, n = 5 seeds")
```

Paired per-seed contrasts against the baseline arm. Same seed => same test patches,
so each row is a paired difference, and at n=5 the credible summary is the sign
consistency (`n_better` / `n_seeds`), not a p-value. `lower_is_better` marks the
confusion rates, where a negative delta is the improvement.

```{r}
#| label: q7-contrasts

read_arch_csv("arch_contrasts.csv") |>
  filter(metric %in% c("macro_f1", "mean_iou") | str_detect(metric, "^conf_")) |>
  mutate(arch = arch_name[arch], baseline = arch_name[baseline],
         across(c(baseline_mean, arch_mean, delta_mean, delta_sd), \(x) round(x, 4)),
         # pandas writes "True"/"False"; read.csv leaves those as character
         across(c(sign_consistent, lower_is_better), as.logical),
         seeds = paste0(n_better, "/", n_seeds)) |>
  select(mode, arch, baseline, metric, baseline_mean, arch_mean,
         delta_mean, delta_sd, seeds, sign_consistent, lower_is_better) |>
  arrange(mode, arch, metric) |>
  knitr::kable(caption = "Paired per-seed deltas vs the baseline architecture")
```

What the extra accuracy costs. Params come from the training journal (exact);
GFLOPs from Lightning's model summary in `train.log`, so they are blank for cells
whose Lightning version did not print that line. `sec_per_epoch` is fit-only wall
clock and is only recorded for v3 runs onwards -- it is a same-GPU comparison, not
a portable benchmark.

```{r}
#| label: q7-cost

read_arch_csv("arch_cost.csv") |>
  mutate(arch = arch_name[arch],
         params_M = round(params / 1e6, 2),
         params_x = round(params_x_baseline, 3)) |>
  select(mode, arch, params_M, params_x, gflops, sec_per_epoch, epochs_run) |>
  arrange(mode, desc(params_M)) |>
  knitr::kable(caption = "Cost per architecture (seed means)")
```

### What do the fusion gates attend to?

`dl_11_export_gates.py` writes one `.npz` per held-out field patch under
`<cell>/seed<k>/gates/`, holding the per-scale softmax gate weights (one channel
per branch, summing to 1 at every pixel, stored float16) plus a
`gate_summary.json` of the per-branch means. Eight patches per cell x 5 seeds x
2 modes, the same held-out patches the cell was scored on.

**Read these as spatial patterns within a branch, not as a cross-branch ranking.**
After gating, `proj` is a 1x1 conv, so the decoder sees `sum_i W_i(f_i * g_i)`:
the gate is a spatial scalar but `W_i` is a learned linear map with its own
magnitude, and GroupNorm equalizes the features, not the projection weights. So a
branch with modest gates and large weights can still dominate. What is
interpretable is (a) where within a scene a branch's weight rises relative to its
*own* average, and (b) whether the gates stay mixed rather than collapsing onto
one branch. Overall branch importance comes from SHAP, not from here.

```{r}
#| label: q8-gate-reader

# .npy / .npz reader in base R. np.savez_compressed writes a deflate ZIP of .npy
# members and dl_11 stores gates as float16, so utils::unzip handles the container
# and the half-float decode below handles the payload -- no numpy/reticulate.
# A 65536-entry lookup turns the uint16 -> double conversion into an index.
HALF_LUT <- local({
  h <- 0:65535
  s <- bitwAnd(bitwShiftR(h, 15L), 1L)
  e <- bitwAnd(bitwShiftR(h, 10L), 0x1FL)
  m <- bitwAnd(h, 0x3FFL)
  v <- ifelse(e == 0L,  2^-14 * (m / 1024),
       ifelse(e == 31L, ifelse(m == 0L, Inf, NaN), 2^(e - 15L) * (1 + m / 1024)))
  ifelse(s == 1L, -v, v)
})

read_npy <- function(path) {
  con <- file(path, "rb"); on.exit(close(con))
  stopifnot(rawToChar(readBin(con, "raw", 6L)[2:6]) == "NUMPY")
  ver  <- as.integer(readBin(con, "raw", 2L))
  hlen <- if (ver[1] >= 2L) readBin(con, "integer", 1L, size = 4L, endian = "little")
          else readBin(con, "integer", 1L, size = 2L, endian = "little", signed = FALSE)
  hdr   <- rawToChar(readBin(con, "raw", hlen))
  descr <- str_match(hdr, "'descr':\\s*'([^']+)'")[, 2]
  dims  <- str_trim(str_split(str_match(hdr, "'shape':\\s*\\(([^)]*)\\)")[, 2], ",")[[1]])
  shape <- as.integer(dims[nzchar(dims)])
  n     <- prod(shape)
  vals  <- switch(descr,
    "<f2" = HALF_LUT[readBin(con, "integer", n, size = 2L, endian = "little", signed = FALSE) + 1L],
    "<f4" = readBin(con, "double", n, size = 4L, endian = "little"),
    "<f8" = readBin(con, "double", n, size = 8L, endian = "little"),
    stop("unsupported .npy dtype: ", descr))
  # numpy is row-major, R is column-major: fill the reversed shape, then permute
  # back so the array indexes as [branch, row, col].
  aperm(array(vals, dim = rev(shape)), rev(seq_along(shape)))
}

read_npz <- function(path, members = NULL) {
  tmp <- file.path(tempdir(), paste0("npz_", basename(path)))
  dir.create(tmp, showWarnings = FALSE, recursive = TRUE)
  on.exit(unlink(tmp, recursive = TRUE))
  files <- unzip(path, exdir = tmp)
  names(files) <- tools::file_path_sans_ext(basename(files))
  if (!is.null(members)) files <- files[intersect(members, names(files))]
  map(files, read_npy)
}

# MOD_CLASS from the patch a gate raster came from, as a [row, col] matrix on the
# same grid as level0. Labels on disk are always the multiclass codes -- the
# binary collapse happens in the dataset -- so binary cells get the same 4 classes.
patch_labels <- function(stem) {
  f <- file.path(patch_dir, paste0(stem, ".tif"))
  if (!file.exists(f)) return(NULL)
  r  <- terra::rast(f)
  lb <- r[[which(names(r) == "MOD_CLASS")]]
  matrix(as.vector(terra::values(lb)), nrow = nrow(lb), ncol = ncol(lb), byrow = TRUE)
}

as_tiles <- function(m) {
  expand_grid(row = seq_len(nrow(m)), col = seq_len(ncol(m))) |>
    mutate(value = as.vector(t(m)), x = col, y = row)
}
```

```{r}
#| label: q8-gate-cells

gate_summaries <- Sys.glob(file.path(arch_dir_base, "*", "*_mbfusion", "seed*",
                                     "gates", "gate_summary.json"))
have_gates <- length(gate_summaries) > 0

if (!have_gates) {
  cat("No gate exports found under", arch_dir_base,
      "\nRun dl_11_export_gates.py on each mbfusion cell and rsync the gates/ dirs back.\n")
} else {
  gate_cells <- tibble(summary_file = gate_summaries) |>
    mutate(js       = map(summary_file, ~ fromJSON(.x, simplifyVector = TRUE)),
           mode     = factor(str_to_title(map_chr(js, "mode")),
                             levels = c("Multiclass", "Binary")),
           seed     = map_int(js, ~ as.integer(.x$seed)),
           branches = map(js, "branches"),
           gate_dir = dirname(summary_file))

  # Branch order is the checkpoint's own gate-channel order, taken from the JSON.
  branch_levels <- gate_cells$branches[[1]]
  branch_label  <- c(terrain = "Terrain", lidar = "Lidar",
                     leafon  = "NAIP leaf-on", leafoff = "NAIP leaf-off")
  branch_color  <- c(Terrain = "#6a4836", Lidar = "#81b14f",
                     `NAIP leaf-on` = "#388C73", `NAIP leaf-off` = "#B07AA1")
  BRANCHES  <- unname(branch_label[branch_levels])
  n_branch  <- length(branch_levels)
  # MOD_CLASS integer codes -> names, per dl_band_config.json
  mod_class <- c(`0` = "EMW", `1` = "FSW", `2` = "SSW", `3` = "UPL")

  cat(nrow(gate_cells), "mbfusion cells with gate exports;",
      n_branch, "branches:", paste(BRANCHES, collapse = ", "), "\n")
}
```

Mean gate weight per branch and scale, from `gate_summary.json`. The dashed line
is the uniform share (`1/n_branch`). **This is provenance, not a ranking** -- read
each branch's own trajectory across scales, not the branches against each other.

```{r}
#| label: q8-gate-means
#| fig-width: 16
#| fig-height: 7
#| eval: !expr exists("gate_cells")

gate_means <- pmap_dfr(list(gate_cells$js, gate_cells$mode, gate_cells$seed),
  function(js, md, sd) {
    imap_dfr(js$mean_gate, function(v, sc)
      tibble(mode = md, seed = sd, scale = sc,
             branch = names(v), mean_gate = as.numeric(unlist(v))))
  }) |>
  mutate(scale  = factor(scale, levels = paste0("level", 0:5)),
         branch = factor(unname(branch_label[branch]), levels = BRANCHES))

gate_means |>
  ggplot(aes(x = scale, y = mean_gate, colour = branch, group = branch)) +
    geom_hline(yintercept = 1 / n_branch, linetype = 2, colour = "grey55") +
    geom_jitter(width = 0.10, alpha = 0.30, size = 2) +
    stat_summary(fun = mean, geom = "line", linewidth = 1.2) +
    stat_summary(fun = mean, geom = "point", size = 4) +
    facet_wrap(~ mode) +
    scale_colour_manual(values = branch_color, name = NULL) +
    scale_y_continuous(labels = label_percent(accuracy = 1)) +
    labs(x = "Encoder scale (level0 = full resolution, level5 = bottleneck)",
         y = "Mean gate weight",
         subtitle = paste0("Seed points jittered; dashed line = uniform share (1/",
                           n_branch, "). Compare a branch across scales, not branches to each other.")) +
    theme_param
```

The collapse check. Per-pixel gate entropy `-sum g log g`, averaged over every
exported pixel: `log(n_branch)` = `r if (exists("n_branch")) round(log(n_branch), 3) else NA`
means a perfectly even blend, 0 means the gate has gone hard onto a single branch.
This is the one number that says whether the fusion is actually fusing.

```{r}
#| label: q8-gate-pixels
#| cache: true
#| eval: !expr exists("gate_cells")

# Walks every .npz once (~80 files, ~30-60 s) and returns both pixel-level
# summaries, so the float16 decode is paid for a single time.
gate_pixels <- pmap_dfr(list(gate_cells$gate_dir, gate_cells$mode, gate_cells$seed),
  function(dir, md, sd) {
    map_dfr(Sys.glob(file.path(dir, "*.npz")), function(f) {
      stem <- tools::file_path_sans_ext(basename(f))
      lv   <- read_npz(f, members = paste0("level", 0:5))

      # (a) per-scale mean gate entropy
      ent <- imap_dfr(lv, function(a, nm) {
        g <- pmax(a, 1e-6)   # float16 stores exact zeros; keep log() finite
        tibble(stat = "entropy", scale = nm,
               entropy = mean(colSums(-g * log(g), dims = 1)))
      })

      # (b) level0 gate conditioned on the true class -- level0 is 256x256, the
      # same grid as the patch, so gate pixels and label pixels line up 1:1.
      labm <- patch_labels(stem)
      cls  <- if (is.null(labm)) NULL else {
        code <- as.vector(t(labm))
        map_dfr(seq_along(branch_levels), function(k)
          tibble(branch = branch_levels[k], code = code, g = as.vector(t(lv$level0[k, , ]))) |>
            filter(code %in% as.integer(names(mod_class))) |>
            group_by(branch, code) |>
            summarise(mean_gate = mean(g), n = n(), .groups = "drop")) |>
          mutate(stat = "byclass")
      }

      bind_rows(ent, cls) |> mutate(mode = md, seed = sd, patch = stem)
    })
  })
```

```{r}
#| label: q8-gate-entropy
#| fig-width: 16
#| fig-height: 7
#| eval: !expr exists("gate_pixels")

gate_pixels |>
  filter(stat == "entropy") |>
  mutate(scale = factor(scale, levels = paste0("level", 0:5))) |>
  ggplot(aes(x = scale, y = entropy)) +
    geom_hline(yintercept = log(n_branch), linetype = 2, colour = "grey55") +
    geom_jitter(aes(colour = factor(seed)), width = 0.12, alpha = 0.45, size = 2) +
    stat_summary(fun = mean, geom = "line", aes(group = 1), linewidth = 1.2) +
    stat_summary(fun = mean, geom = "point", size = 4) +
    facet_wrap(~ mode) +
    scale_colour_grey(start = 0.1, end = 0.7, name = "Seed") +
    scale_y_continuous(limits = c(0, log(n_branch) * 1.05)) +
    labs(x = "Encoder scale (level0 = full resolution, level5 = bottleneck)",
         y = "Mean per-pixel gate entropy",
         subtitle = paste0("Dashed line = log(", n_branch, ") = ", round(log(n_branch), 3),
                           ", a perfectly even blend. 0 = the gate has gone hard onto one branch.",
                           "\nPoints are patches, coloured by seed; line is the mean.")) +
    theme_param
```

Where each branch is up-weighted, by the true class of the pixel. Boxes are the
**deviation from that branch's own mean gate** over the same pixels, in percentage
points, so every box is a legitimate within-branch spatial comparison. Positive =
the model leans on that modality more than usual over that class.

Each panel uses the class set its own models were trained and scored on: the
multiclass row keeps EMW/FSW/SSW/UPL, the binary row shows WET/UPL. The patches on
disk always carry the four `MOD_CLASS` codes -- the collapse happens in the dataset
-- so the binary row is built by pooling the EMW/FSW/SSW **pixels** into WET,
weighted by pixel count, before taking the deviation. Relabelling the three classes
without pooling would draw three separate boxes all called "WET" and would not be
the quantity the binary model actually sees.

```{r}
#| label: q8-gate-class
#| fig-width: 16
#| fig-height: 8
#| eval: !expr exists("gate_pixels")

gate_class <- gate_pixels |>
  filter(stat == "byclass") |>
  mutate(class = unname(mod_class[as.character(code)]),
         # binary cells are scored on WET/UPL, so pool the wetland pixels rather
         # than relabelling three classes "WET". `n` is the pixel count, so the
         # weighted mean below does the patch-average and the class-collapse at once.
         class = if_else(mode == "Binary" & class %in% binary_wet, "WET", class)) |>
  group_by(mode, seed, branch, class) |>
  summarise(mean_gate = weighted.mean(mean_gate, n), n = sum(n), .groups = "drop") |>
  group_by(mode, seed, branch) |>
  mutate(delta = mean_gate - weighted.mean(mean_gate, n)) |>   # vs the branch's own mean
  ungroup() |>
  mutate(class  = factor(class, levels = class_order),
         branch = factor(unname(branch_label[branch]), levels = BRANCHES))

# facet_grid shares the x scale down a column, so the 4-class and 2-class rows
# have to be separate plots stacked by patchwork.
gate_class_fig <- function(m) {
  gate_class |>
    filter(mode == m) |>
    ggplot(aes(x = class, y = delta, fill = class)) +
      geom_hline(yintercept = 0, colour = "grey40") +
      box_jitter(point_size = 3) +
      facet_grid(. ~ branch) +
      scale_fill_manual(values = class_color, guide = "none") +
      scale_y_continuous(labels = label_percent(accuracy = 1)) +
      labs(x = NULL, y = "Gate weight minus\nthe branch's own mean", title = m) +
      theme_param
}

gate_class_fig("Multiclass") / gate_class_fig("Binary") +
  plot_annotation(
    caption = "Points are seeds. Positive = that modality is up-weighted over that class.",
    theme = theme(plot.caption = element_text(size = 14, colour = "grey35", hjust = 0)))
```

Finally the maps themselves, for one seed: the field labels beside each branch's
full-resolution (level0) gate, standardized **within branch** to SD units so the
panels show spatial pattern rather than the branch's overall level.

```{r}
#| label: q8-gate-maps
#| fig-width: 16
#| fig-height: 9
#| eval: !expr exists("gate_cells")

MAP_MODE <- "Multiclass"   # which arm's cells to draw
MAP_SEED <- 0
MAP_N    <- 2              # patches to draw (8 are exported per cell)

theme_map <- theme_param +
  theme(axis.text = element_blank(), axis.ticks = element_blank(),
        panel.grid = element_blank(),
        plot.title = element_text(size = rel(0.75), face = "bold"),
        plot.subtitle = element_text(size = rel(0.6), colour = "grey35"))

gate_patch_row <- function(gate_dir, stem, level = "level0") {
  a    <- read_npz(file.path(gate_dir, paste0(stem, ".npz")), members = level)[[level]]
  labm <- patch_labels(stem)

  p_lab <- as_tiles(labm) |>
    mutate(class = factor(mod_class[as.character(value)], levels = class_order)) |>
    ggplot(aes(x, y, fill = class)) +
      geom_raster() +
      # drop = FALSE keeps one shared legend across rows with different classes
      scale_fill_manual(values = class_color, na.value = "grey90", name = NULL,
                        na.translate = FALSE, drop = FALSE,
                        limits = names(mod_class) |> (\(k) unname(mod_class[k]))()) +
      scale_y_reverse(expand = c(0, 0)) + scale_x_continuous(expand = c(0, 0)) +
      coord_fixed() +
      labs(x = NULL, y = NULL, title = "Field labels",
           subtitle = str_trunc(stem, 46)) +
      theme_map

  p_gate <- imap_dfr(setNames(seq_along(branch_levels), branch_levels),
                     function(k, bn) as_tiles(a[k, , ]) |> mutate(branch = bn)) |>
    group_by(branch) |>
    mutate(z = (value - mean(value)) / sd(value)) |>
    ungroup() |>
    mutate(branch = factor(unname(branch_label[branch]), levels = BRANCHES)) |>
    ggplot(aes(x, y, fill = z)) +
      geom_raster() +
      facet_wrap(~ branch, nrow = 1) +
      scale_fill_gradient2(low = "#2166ac", mid = "grey96", high = "#b2182b",
                           midpoint = 0, limits = c(-2.5, 2.5), oob = squish,
                           name = "SD from the\nbranch's own\nmean gate") +
      scale_y_reverse(expand = c(0, 0)) + scale_x_continuous(expand = c(0, 0)) +
      coord_fixed() +
      labs(x = NULL, y = NULL) +
      theme_map

  p_lab + p_gate + plot_layout(widths = c(1, 4))
}

map_cell <- gate_cells |> filter(mode == MAP_MODE, seed == MAP_SEED)
if (nrow(map_cell) == 0) {
  cat("No mbfusion cell for", MAP_MODE, "seed", MAP_SEED, "\n")
} else {
  stems <- tools::file_path_sans_ext(basename(
    Sys.glob(file.path(map_cell$gate_dir[1], "*.npz"))))
  stems <- stems[file.exists(file.path(patch_dir, paste0(stems, ".tif")))]
  # spread the picks over the exported patches rather than taking the first N,
  # which would all come from the same HUC12
  picks <- stems[unique(round(seq(1, length(stems), length.out = min(MAP_N, length(stems)))))]
  wrap_plots(map(picks, ~ gate_patch_row(map_cell$gate_dir[1], .x)), ncol = 1) +
    plot_layout(guides = "collect")
}
```
