---
title: "Factorial Experiment Visualization (v2)"
author: "Anthony Stewart"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
    fig-width: 16
    fig-height: 9
execute:
  warning: false
  message: false
---

Walkthrough of the v2 factorial experiment results (8 configs x 5 seeds x
{multiclass, binary}) plus the follow-on studies (patch-count learning curve,
UNet3+ comparison). Inputs are the small CSV/JSON aggregation outputs only:

- `dl_08_aggregate_factorial.py` -> `Models/factorial_results_v2/<mode>/analysis/`
- `dl_08b_aggregate_patchcurve.py` -> `Models/results_patchcurve_v2/<mode>/analysis/` and `Models/results_arch_v2/<mode>/analysis/`
- `dl_09b_shap_geomorph_export.py` -> `Models/factorial_results_v2/<mode>/analysis/shap_geomorph_channels.csv`

```{r}
#| label: setup

library(tidyverse)
library(patchwork)
library(jsonlite)
library(colorspace)
library(scales)
```


```{r}
#| label: factorial-data

multi_dir  <- "../Models/factorial_results_v2/multiclass"
binary_dir <- "../Models/factorial_results_v2/binary"

multi_fact_data <- read.csv(file.path(multi_dir, "analysis/factorial_long.csv"))
glimpse(multi_fact_data)

binary_fact_data <- read.csv(file.path(binary_dir, "analysis/factorial_long.csv"))
glimpse(binary_fact_data)
```

```{r}
#| label: results-roots

# Root holding the architecture-comparison analysis/ dirs -- i.e. the LAST
# --arch-dir passed to dl_08b_aggregate_patchcurve.py, which is where it writes.
# v2: results_arch_v2 (two arms). For v3 point this at the fusion root, which
# also holds the gate exports read by the last section:
#   arch_dir_base <- "../Models/results_arch_fusion_v3"
arch_dir_base <- "../Models/results_arch_v2"
```

```{r}
#| label: data-organizing

config_order <- c("fld_nolidar_leafon", "fld_nolidar_leafoff",  "fld_chmret_leafon", "fld_chmret_leafoff",
                 "nwi_chmret_leafoff", "nwiextra_chmret_leafoff", "nwifield_chmret_leafoff", "flddeg_chmret_leafoff")

config_label <- c("fld_nolidar_leafon" = "Baseline",
                  "fld_nolidar_leafoff" = "With\nLeaf-Off",
                  "fld_chmret_leafon" = "With\nLidar",
                  "fld_chmret_leafoff" = "Full\nModel",
                  "nwi_chmret_leafoff" = "NWI\nData",
                  "nwiextra_chmret_leafoff" = "NWI\nx2",
                  "nwifield_chmret_leafoff" = "NWI\n+\nField",
                  "flddeg_chmret_leafoff" = "Degraded\nField")

# single-line variant for heatmap y-axes (stacked labels overlap there)
config_label_1line <- gsub("\n ", ", ", config_label)

class_order <- c("EMW", "FSW", "SSW", "WET", "UPL")
class_color <- c("FSW" = "#81b14f", "SSW" = "#c4bd8b", "EMW" = "#8faab3", "UPL" = "#6a4836", "WET" = "#388C73")

metric_label <- c("precision" = "Precision", "recall" = "Recall", "f1" = "F1", "iou" = "IoU",
                  "macro_f1" = "Macro F1", "mean_iou" = "Mean IoU", "overall_accuracy" = "Overall accuracy")

organize <- function(df) {
  df |>
    mutate(config = factor(config, levels = config_order),
           class  = factor(class,  levels = class_order))
}
multi_fact_data  <- organize(multi_fact_data)
binary_fact_data <- organize(binary_fact_data)

# Both modes in one frame for faceted figures
fact_data <- bind_rows(
  multi_fact_data  |> mutate(mode = "Multiclass"),
  binary_fact_data |> mutate(mode = "Binary")
) |> mutate(mode = factor(mode, levels = c("Multiclass", "Binary")))

theme_param <- theme(
  text = element_text(size = 18),
  legend.position = 'right',
  legend.text = element_text(size = rel(0.70), hjust = 0.5),
  panel.background = element_blank(),
  panel.grid.major = element_line(colour = "grey50", linewidth = 0.2),
  axis.ticks = element_blank(),
  axis.text = element_text(size = rel(0.9)),
  axis.title = element_text(size = rel(1)),
  strip.background = element_blank(),
  strip.text = element_text(size = rel(0.9), face = "bold")
)

theme_heatmap <- theme_param + theme(panel.grid.major = element_blank())

# Boxplot + jittered seed points, shared across the metric figures
box_jitter <- function(point_size = 5) {
  list(
    geom_boxplot(outlier.shape = NA, position = position_dodge(width = 0.75)),
    geom_jitter(position = position_jitterdodge(jitter.width = 0.15, dodge.width = 0.75),
                size = point_size, shape = 21, colour = "grey15")
  )
}

# Single-hue sequential fill for the heatmaps (magnitude = light -> dark)
seq_fill <- function(...) {
  scale_fill_gradient(low = "#f2f7f5", high = "#14523f", na.value = "grey93", ...)
}
```

### Conceptual figure or matrix for the factorial experiment design

Structure

- Sized for a slide: 16:9, few elements, large type, plain language (no "ablation")
- Two tables sharing one model: table A varies the inputs with the field labels held constant,
  table B varies the training labels with the full input set held constant. The orange
  `fld_chmret_leafoff` model is the single model both tables have in common
- Only the inputs that actually *vary* get columns. Terrain, geomorphon and leaf-on NAIP are in
  every model, so they are stated once in the subtitle instead of drawn as always-filled columns
- Built from `fact_data` itself (label / lidar / spectral / in_channels are columns of
  `factorial_long.csv`), so the figure cannot drift out of sync with what actually ran

```{r}
#| label: design-matrix
#| fig-width: 16
#| fig-height: 9

reference <- "fld_chmret_leafoff"
ref_col   <- "#d95f02"

design_cells <- fact_data |>
  distinct(config, label, lidar, spectral, in_channels) |>
  mutate(config = as.character(config))

# ---- Table A: inputs vary, field labels held constant ------------------------
col_lidar    <- "Lidar\n(CHM + return fractions)"
col_leafoff  <- "NAIP\nleaf-off"
col_channels <- "Input\nchannels"
a_cols <- c(col_lidar, col_leafoff, col_channels)

a_row <- c("fld_nolidar_leafon"  = "Baseline",
           "fld_nolidar_leafoff" = "+ Leaf-off",
           "fld_chmret_leafon"   = "+ Lidar",
           "fld_chmret_leafoff"  = "Full inputs")

a_cells <- design_cells |>
  filter(label == "fld") |>
  mutate(config = factor(config, levels = rev(config_order)))

# "in"/"out" drive the fill; "meta" is the un-filled channel-count column
a_tiles <- bind_rows(
  a_cells |>
    mutate("{col_lidar}"   := if_else(lidar == "chmret", "in", "out"),
           "{col_leafoff}" := if_else(spectral == "leafoff", "in", "out")) |>
    pivot_longer(all_of(c(col_lidar, col_leafoff)), names_to = "input", values_to = "state"),
  a_cells |> mutate(input = col_channels, state = "meta")
) |>
  mutate(input = factor(input, levels = a_cols))

p_a <- ggplot(a_tiles, aes(x = input, y = config)) +
  geom_tile(aes(fill = state), colour = "white", linewidth = 1.2) +
  geom_tile(data = filter(a_tiles, config == reference, state != "meta"),
            fill = NA, colour = ref_col, linewidth = 1.8) +
  geom_text(data = filter(a_tiles, state == "meta"), aes(label = in_channels),
            colour = "grey15", size = 6.5, fontface = "bold") +
  scale_fill_manual(values = c("in" = "#14523f", "out" = "grey88", "meta" = "white"),
                    breaks = c("in", "out"), labels = c("In the model", "Left out"),
                    name = NULL, na.value = "white") +
  scale_x_discrete(limits = a_cols, position = "top") +
  scale_y_discrete(labels = a_row) +
  labs(x = NULL, y = NULL, title = "A · Field labels held constant — do lidar and leaf-off imagery help?")

# ---- Table B: training labels vary, full input set held constant --------------
b_head <- c("fld" = "Field", "nwi" = "NWI", "nwiextra" = "NWI x2",
            "nwifield" = "NWI + Field", "flddeg" = "Degraded field")

b_desc <- c("fld"      = "Field- and GIS-\nverified polygons",
            "nwi"      = "NWI polygons, paired\nto the field patches",
            "nwiextra" = "NWI polygons,\n~2x the patches",
            "nwifield" = "Field plus extra\nNWI patches pooled",
            "flddeg"   = "Field labels thinned to\nNWI wetland prevalence")

b_cells <- design_cells |>
  filter(lidar == "chmret", spectral == "leafoff") |>
  mutate(label = factor(label, levels = names(b_head)),
         row   = paste0("Full inputs\n(", in_channels, " channels)"))

p_b <- ggplot(b_cells, aes(x = label, y = row)) +
  geom_tile(fill = "#14523f", colour = "white", linewidth = 1.2) +
  geom_tile(data = filter(b_cells, config == reference),
            fill = NA, colour = ref_col, linewidth = 1.8) +
  geom_text(aes(label = b_desc[as.character(label)]),
            colour = "white", size = 4.6, lineheight = 0.95) +
  scale_x_discrete(labels = b_head, position = "top") +
  labs(x = NULL, y = NULL, title = "B · Full input set held constant — which training labels?")

n_seeds <- n_distinct(fact_data$seed)

theme_design <- theme_heatmap +
  theme(text = element_text(size = 22),
        legend.position = "bottom",
        legend.text = element_text(size = rel(0.8), hjust = 0),
        axis.text.x = element_text(size = rel(0.7), lineheight = 0.9),
        axis.text.y = element_text(size = rel(0.8), hjust = 1, lineheight = 0.9),
        plot.title = element_text(size = rel(0.8), face = "bold"),
        plot.title.position = "plot")

(p_a + theme_design) / (p_b + theme_design) +
  plot_layout(heights = c(4, 1.6), guides = "collect") +
  plot_annotation(
    title = "Factorial design: 8 model configurations",
    subtitle = paste0("Every model includes terrain, geomorphon and leaf-on NAIP, and is trained at ",
                      n_seeds, " seeds x {multiclass, binary}.\n",
                      "Every model is evaluated against field-verified labels, whatever it trained on."),
    caption = paste0("Orange = the reference model (field labels + lidar + leaf-off): the one model both ",
                     "tables share, and the\nstarting point for the follow-on studies — patch-count curve (Q6) ",
                     "and UNet3+ (Q7)."),
    theme = theme(plot.title = element_text(size = 26, face = "bold"),
                  plot.subtitle = element_text(size = 15, colour = "grey30", lineheight = 1.1),
                  plot.caption = element_text(size = 13, colour = "grey35", hjust = 0, lineheight = 1.1),
                  plot.title.position = "plot",
                  plot.caption.position = "plot",
                  legend.position = "bottom",
                  plot.margin = margin(12, 18, 12, 12))
  )
```


### Does leaf off imagery or lidar-derived metrics improve field-based models?

Structure

- Just using the the field-verified dataset models
- Model config on the x-axis, model evaluation metric on the y-axis
- box and whiskers plots representing the model measure spread, points for the individual model measures
- Precision and recall share a panel: full class color = precision, lighter shade = recall

```{r}
#| label: q1-precision-recall
#| fig-width: 8
#| fig-height: 14
#| layout-ncol: 2

pr_data <- fact_data |>
  filter(label == "fld") |>
  pivot_longer(c(precision, recall), names_to = "metric", values_to = "value") |>
  mutate(metric = factor(metric_label[metric], levels = c("Precision", "Recall")))

pr_plot <- function(which_mode, which_metric) {
  pr_data |>
    filter(mode == which_mode, metric == which_metric) |>
    ggplot(aes(x = config, y = value, fill = class)) +
      facet_wrap(~ class, scales = "free_y", ncol = 1) +
      box_jitter(point_size = 3.5) +
      scale_fill_manual(values = class_color) +
      scale_x_discrete(labels = config_label) +
      labs(x = NULL, y = which_metric, title = which_mode) +
      theme_param
}

pr_plot("Multiclass", "Precision")

pr_plot("Multiclass", "Recall")

pr_plot("Binary", "Precision")

pr_plot("Binary", "Recall")
```

```{r}
#| label: q1-f1-iou
#| fig-width: 8
#| fig-height: 14

fact_data |>
  filter(label == "fld", mode == "Multiclass") |>
  ggplot(aes(x = config, y = f1, fill = class)) +
    box_jitter() +
    facet_wrap(~ class, scales = "free_y", ncol = 1) +
    scale_fill_manual(values = class_color) +
    scale_x_discrete(labels = config_label) +
    scale_y_continuous(labels = scales::label_number(accuracy = 0.01)) +
    labs(x = NULL, y = "F1") +
    # ylim(0.0, 1.0) +
    theme_param

fact_data |>
  filter(label == "fld", mode == "Multiclass") |>
  ggplot(aes(x = config, y = iou, fill = class)) +
    box_jitter() +
    facet_wrap(~ class, scales = "free_y", ncol = 1) +
    scale_fill_manual(values = class_color) +
    scale_x_discrete(labels = config_label) +
    scale_y_continuous(labels = scales::label_number(accuracy = 0.01)) +
    labs(x = NULL, y = "IoU") +
    theme_param

  fact_data |>
  filter(label == "fld", mode == "Binary") |>
  ggplot(aes(x = config, y = f1, fill = class)) +
    box_jitter() +
    facet_wrap(~ class, scales = "free_y", ncol = 1) +
    scale_fill_manual(values = class_color) +
    scale_x_discrete(labels = config_label) +
    scale_y_continuous(labels = scales::label_number(accuracy = 0.01)) +
    labs(x = NULL, y = "F1") +
    theme_param

fact_data |>
  filter(label == "fld", mode == "Binary") |>
  ggplot(aes(x = config, y = iou, fill = class)) +
    box_jitter() +
    facet_wrap(~ class, scales = "free_y", ncol = 1) +
    scale_fill_manual(values = class_color) +
    scale_x_discrete(labels = config_label) +
    scale_y_continuous(labels = scales::label_number(accuracy = 0.01)) +
    labs(x = NULL, y = "IoU") +
    theme_param
```


### How do NWI data labels compare to field and GIS-derived data?

#### Also does degrading the field data hinder its performance?

- All label sources on the full feature set (`chmret_leafoff`), evaluated against field labels

```{r}
#| label: q2-nwi-f1-iou
#| fig-width: 24
#| fig-height: 6

config_label_nwi <- config_label
config_label_nwi[4] <- "Full\nModel\nField+GIS"

fact_data |>
  filter(featureset == "chmret_leafoff", mode == "Multiclass") |>
  ggplot(aes(x = config, y = f1, fill = class)) +
    box_jitter() +
    facet_wrap(mode ~ class, scales = "free_y", ncol = 4) +
    scale_fill_manual(values = class_color) +
    scale_x_discrete(labels = config_label_nwi) +
    scale_y_continuous(labels = scales::label_number(accuracy = 0.01)) +
    labs(x = NULL, y = "F1") +
    theme_param

fact_data |>
  filter(featureset == "chmret_leafoff", mode == "Binary") |>
  ggplot(aes(x = config, y = f1, fill = class)) +
    box_jitter() +
    facet_wrap(mode ~ class, scales = "free_y", ncol = 4) +
    scale_fill_manual(values = class_color) +
    scale_x_discrete(labels = config_label) +
    scale_y_continuous(labels = scales::label_number(accuracy = 0.01)) +
    labs(x = NULL, y = "F1") +
    theme_param

# fact_data |>
#   filter(featureset == "chmret_leafoff", mode == "Multiclass") |>
#   ggplot(aes(x = config, y = iou, fill = class)) +
#     box_jitter() +
#     facet_wrap(mode ~ class, scales = "free_y", ncol = 4) +
#     scale_fill_manual(values = class_color) +
#     scale_x_discrete(labels = config_label) +
#     scale_y_continuous(labels = scales::label_number(accuracy = 0.01)) +
#     labs(x = NULL, y = "IoU") +
#     theme_param

# fact_data |>
#   filter(featureset == "chmret_leafoff", mode == "Binary") |>
#   ggplot(aes(x = config, y = iou, fill = class)) +
#     box_jitter() +
#     facet_wrap(mode ~ class, scales = "free_y", ncol = 4) +
#     scale_fill_manual(values = class_color) +
#     scale_x_discrete(labels = config_label) +
#     scale_y_continuous(labels = scales::label_number(accuracy = 0.01)) +
#     labs(x = NULL, y = "IoU") +
#     theme_param
```

Model-level summary (macro metrics are repeated on every class row of
`factorial_long.csv`, so collapse to one row per model first):

```{r}
#| label: q2-macro
#| fig-width: 20
#| fig-height: 8

fact_data |>
  filter(featureset == "chmret_leafoff", mode == "Multiclass") |>
  distinct(mode, config, seed, macro_f1) |>
  pivot_longer(c(macro_f1), names_to = "metric", values_to = "value") |>
  mutate(metric = metric_label[metric]) |>
  ggplot(aes(x = config, y = value, fill = metric)) +
    box_jitter(point_size = 4) +
    facet_grid(. ~ mode) +
    scale_fill_manual(values = c("Macro F1" = "#388C73", "Mean IoU" = lighten("#388C73", 0.45)), name = NULL) +
    scale_x_discrete(labels = config_label) +
    labs(x = NULL, y = NULL) +
    theme_param
  
fact_data |>
  filter(featureset == "chmret_leafoff", mode == "Multiclass") |>
  distinct(mode, config, seed, mean_iou) |>
  pivot_longer(c(mean_iou), names_to = "metric", values_to = "value") |>
  mutate(metric = metric_label[metric]) |>
  ggplot(aes(x = config, y = value, fill = metric)) +
    box_jitter(point_size = 4) +
    facet_grid(. ~ mode) +
    scale_fill_manual(values = c("Macro F1" = "#388C73", "Mean IoU" = lighten("#388C73", 0.45)), name = NULL) +
    scale_x_discrete(labels = config_label) +
    labs(x = NULL, y = NULL) +
    theme_param

fact_data |>
  filter(featureset == "chmret_leafoff", mode == "Binary") |>
  distinct(mode, config, seed, macro_f1) |>
  pivot_longer(c(macro_f1), names_to = "metric", values_to = "value") |>
  mutate(metric = metric_label[metric]) |>
  ggplot(aes(x = config, y = value, fill = metric)) +
    box_jitter(point_size = 4) +
    facet_grid(. ~ mode) +
    scale_fill_manual(values = c("Macro F1" = "#388C73", "Mean IoU" = lighten("#388C73", 0.45)), name = NULL) +
    scale_x_discrete(labels = config_label) +
    labs(x = NULL, y = NULL) +
    theme_param
  
fact_data |>
  filter(featureset == "chmret_leafoff", mode == "Binary") |>
  distinct(mode, config, seed, mean_iou) |>
  pivot_longer(c(mean_iou), names_to = "metric", values_to = "value") |>
  mutate(metric = metric_label[metric]) |>
  ggplot(aes(x = config, y = value, fill = metric)) +
    box_jitter(point_size = 4) +
    facet_grid(. ~ mode) +
    scale_fill_manual(values = c("Macro F1" = "#388C73", "Mean IoU" = lighten("#388C73", 0.45)), name = NULL) +
    scale_x_discrete(labels = config_label) +
    labs(x = NULL, y = NULL) +
    theme_param
```


### How do field and NWI models compare across each other

- Heat maps of model metrics: config (rows) x class (cols), one panel per metric.

```{r}
#| label: q3-metric-heatmap
#| fig-width: 18
#| fig-height: 14

metric_heat <- fact_data |>
  pivot_longer(c(precision, recall, f1, iou), names_to = "metric", values_to = "value") |>
  mutate(metric = factor(metric_label[metric], levels = c("Precision", "Recall", "F1", "IoU"))) |>
  group_by(mode, config, class, metric) |>
  summarise(value = mean(value), .groups = "drop")

heat_panel <- function(which_mode) {
  metric_heat |>
    filter(mode == which_mode) |>
    ggplot(aes(x = class, y = fct_rev(config), fill = value)) +
      geom_tile(colour = "white", linewidth = 1) +
      geom_text(aes(label = number(value, accuracy = 0.01),
                    colour = value > 0.55), size = 4.5, show.legend = FALSE) +
      facet_grid(. ~ metric) +
      scale_colour_manual(values = c(`TRUE` = "white", `FALSE` = "grey20")) +
      seq_fill(limits = c(0, 1), name = "Seed-mean\nscore") +
      scale_y_discrete(labels = config_label_1line) +
      labs(x = NULL, y = NULL, title = which_mode) +
      theme_heatmap
}

heat_panel("Multiclass") / heat_panel("Binary") +
  plot_layout(guides = "collect", heights = c(1, 1))
```


### How accurate are the classes via confusion matrices?

- binary and multiclass model confusion matrices for each configuration
- color and fill by relative % but include a parentheses with the number of pixels e.g. 70% (3,400,000)
- rows (true class) sum to 100%, so the diagonal reads as recall; counts are seed-mean pixels

```{r}
#| label: q4-confusion-load

read_confusion <- function(mode_dir) {
  map_dfr(config_order, function(cfg) {
    f <- file.path(mode_dir, "analysis/confusion_mean", paste0(cfg, ".csv"))
    if (!file.exists(f)) return(NULL)
    read.csv(f, row.names = 1, check.names = FALSE) |>
      rownames_to_column("true") |>
      pivot_longer(-true, names_to = "pred", values_to = "n") |>
      mutate(config = factor(cfg, levels = config_order))
  }) |>
    group_by(config, true) |>
    mutate(prop = n / sum(n)) |>
    ungroup() |>
    mutate(true = factor(true, levels = rev(class_order)),
           pred = factor(pred, levels = class_order),
           cell_label = paste0(percent(prop, accuracy = 1), "\n(",
                               number(n, accuracy = 0.1, scale_cut = cut_short_scale()), ")"))
}

confusion_plot <- function(mode_dir, title, text_size = 4.5) {
  read_confusion(mode_dir) |>
    ggplot(aes(x = pred, y = true, fill = prop)) +
      geom_tile(colour = "white", linewidth = 1) +
      geom_text(aes(label = cell_label, colour = prop > 0.55),
                size = text_size, lineheight = 0.9, show.legend = FALSE) +
      facet_wrap(~ config, ncol = 4, labeller = labeller(config = config_label)) +
      scale_colour_manual(values = c(`TRUE` = "white", `FALSE` = "grey20")) +
      seq_fill(limits = c(0, 1), labels = percent, name = "% of\ntrue class") +
      labs(x = "Predicted", y = "True", title = title) +
      theme_heatmap
}
```

```{r}
#| label: q4-confusion-multi
#| fig-width: 20
#| fig-height: 11

confusion_plot(multi_dir, "Multiclass")
```

```{r}
#| label: q4-confusion-binary
#| fig-width: 20
#| fig-height: 8

confusion_plot(binary_dir, "Binary", text_size = 5.5)
```


### What were the important features in each model configuration according to SHAP?

- Model heatmap with cross-model comparison: each model normalized to its own total importance, so bands are comparable across configs. Bands a config doesn't use are blank.
- Uses the PER-CHANNEL MEAN aggregation (not the raw sum) so the one-hot Geomorph_local band is not inflated 10x relative to single-channel bands.
- Also show within the Geomorphons category a heatmap of the different categories across models

```{r}
#| label: q5-shap-load

band_order <- c("DEM", "slope_local", "Geomorph_local", "flowacc", "twi",
                "CHM", "pct_below_1m", "pct_1m_to_5m", "pct_above_5m",
                "r", "g", "b", "nir", "r_lo", "g_lo", "b_lo", "nir_lo")

band_label <- c("DEM" = "DEM", "slope_local" = "Slope", "Geomorph_local" = "Geomorphons",
                "flowacc" = "Flow accum.", "twi" = "TWI",
                "CHM" = "CHM", "pct_below_1m" = "% returns < 1 m",
                "pct_1m_to_5m" = "% returns 1-5 m", "pct_above_5m" = "% returns > 5 m",
                "r" = "Red (leaf-on)", "g" = "Green (leaf-on)", "b" = "Blue (leaf-on)", "nir" = "NIR (leaf-on)",
                "r_lo" = "Red (leaf-off)", "g_lo" = "Green (leaf-off)", "b_lo" = "Blue (leaf-off)", "nir_lo" = "NIR (leaf-off)")

read_shap_mode <- function(mode_dir) {
  files <- Sys.glob(file.path(mode_dir, "*", "seed*", "shap", "*_shap_importance.json"))
  map_dfr(files, function(f) {
    j <- fromJSON(f)
    tibble(
      config = basename(dirname(dirname(dirname(f)))),
      seed   = parse_number(str_extract(f, "seed\\d+")),
      band   = names(j$importance_overall_per_channel),
      value  = unlist(j$importance_overall_per_channel)
    )
  })
}

shap_share <- bind_rows(
  read_shap_mode(multi_dir)  |> mutate(mode = "Multiclass"),
  read_shap_mode(binary_dir) |> mutate(mode = "Binary")
) |>
  group_by(mode, config, band) |>
  summarise(value = mean(value), .groups = "drop_last") |>   # mean over seeds
  mutate(share = value / sum(value)) |>                      # normalize per model
  ungroup() |>
  complete(mode, config = config_order, band = band_order) |>  # blanks for unused bands
  mutate(mode   = factor(mode, levels = c("Multiclass", "Binary")),
         config = factor(config, levels = config_order),
         band   = factor(band, levels = rev(band_order)))
```

```{r}
#| label: q5-shap-heatmap
#| fig-width: 16
#| fig-height: 18

shap_share |>
  ggplot(aes(x = config, y = band, fill = share)) +
    geom_tile(colour = "white", linewidth = 1) +
    geom_text(aes(label = percent(share, accuracy = 1), colour = share > 0.10),
              size = 4, na.rm = TRUE, show.legend = FALSE) +
    facet_wrap(~ mode, ncol = 1) +
    scale_colour_manual(values = c(`TRUE` = "white", `FALSE` = "grey20")) +
    seq_fill(labels = percent, name = "Share of model's\ntotal |SHAP|\n(per-channel mean)") +
    scale_x_discrete(labels = config_label) +
    scale_y_discrete(labels = band_label) +
    labs(x = NULL, y = NULL) +
    theme_heatmap
```

Within the Geomorphons band: which landform channels carry the importance
(share of the band's total, seed-mean; forms follow the `rgeomorphon`/GRASS
value order 1-10):

```{r}
#| label: q5-shap-geomorph
#| fig-width: 16
#| fig-height: 12

geomorph_forms <- c("flat", "peak", "ridge", "shoulder", "spur",
                    "slope", "hollow", "footslope", "valley", "pit")

geomorph_shap <- bind_rows(
  read.csv(file.path(multi_dir,  "analysis/shap_geomorph_channels.csv")) |> mutate(mode = "Multiclass"),
  read.csv(file.path(binary_dir, "analysis/shap_geomorph_channels.csv")) |> mutate(mode = "Binary")
) |>
  filter(class == "overall") |>
  group_by(mode, config, geomorphon_form) |>
  summarise(value = mean(mean_abs_shap), .groups = "drop_last") |>
  mutate(share = value / sum(value)) |>
  ungroup() |>
  mutate(mode   = factor(mode, levels = c("Multiclass", "Binary")),
         config = factor(config, levels = config_order),
         geomorphon_form = factor(geomorphon_form, levels = rev(geomorph_forms)))

geomorph_shap |>
  ggplot(aes(x = config, y = geomorphon_form, fill = share)) +
    geom_tile(colour = "white", linewidth = 1) +
    geom_text(aes(label = percent(share, accuracy = 1), colour = share > 0.20),
              size = 4, show.legend = FALSE) +
    facet_wrap(~ mode, ncol = 1) +
    scale_colour_manual(values = c(`TRUE` = "white", `FALSE` = "grey20")) +
    seq_fill(labels = percent, name = "Share of\nGeomorphons\nband |SHAP|") +
    scale_x_discrete(labels = config_label) +
    labs(x = NULL, y = NULL) +
    theme_heatmap
```


### Does the number of patches matter for model accuracy

- F1, IoU, Precision, Recall graphs across the patch numbers input into the model
- Include binary and multiclass stacked graphs
- x-axis is the REALIZED training-set size (from `training_log.json`), points are seeds, lines are seed means
- The `n500` level is dropped: the field-anchored split leaves only ~440-510 training patches, so `n500` realizes essentially the same size as `full` and the two levels overlap.

```{r}
#| label: q6-patchcurve
#| fig-width: 16
#| fig-height: 16

patchcurve <- bind_rows(
  read.csv("../Models/results_patchcurve_v2/multiclass/analysis/patchcurve_long.csv") |> mutate(mode = "Multiclass"),
  read.csv("../Models/results_patchcurve_v2/binary/analysis/patchcurve_long.csv")     |> mutate(mode = "Binary")
) |>
  filter(level != "500") |>   # n500 == full given the ~440-510 patch pool cap
  mutate(mode = factor(mode, levels = c("Multiclass", "Binary")))

pc_class <- patchcurve |>
  pivot_longer(matches("^(iou|f1|precision|recall)_"),
               names_pattern = "^(iou|f1|precision|recall)_(.+)$",
               names_to = c("metric", "class"), values_to = "value") |>
  filter(!is.na(value)) |>   # bind_rows pads the other mode's class columns with NA
  mutate(metric = factor(metric_label[metric], levels = c("Precision", "Recall", "F1", "IoU")),
         class  = factor(class, levels = class_order))

# the realized train count varies a little across seeds within a level,
# so average x and y per level for the mean line
pc_class_mean <- pc_class |>
  group_by(mode, metric, class, level) |>
  summarise(train_patches = mean(train_patches), value = mean(value), .groups = "drop")

pc_class |>
  ggplot(aes(x = train_patches, y = value, colour = class)) +
    geom_point(size = 3, alpha = 0.4) +
    geom_line(data = pc_class_mean, linewidth = 1.2) +
    geom_point(data = pc_class_mean, size = 4) +
    facet_grid(metric ~ mode) +
    scale_colour_manual(values = class_color) +
    labs(x = "Realized training patches", y = NULL) +
    theme_param
```

Model-level curve (macro metrics):

```{r}
#| label: q6-patchcurve-macro
#| fig-width: 16
#| fig-height: 8

pc_macro <- patchcurve |>
  pivot_longer(c(macro_f1, mean_iou, overall_accuracy), names_to = "metric", values_to = "value") |>
  mutate(metric = factor(metric_label[metric], levels = c("Macro F1", "Mean IoU", "Overall accuracy")))

pc_macro_mean <- pc_macro |>
  group_by(mode, metric, level) |>
  summarise(train_patches = mean(train_patches), value = mean(value), .groups = "drop")

pc_macro |>
  ggplot(aes(x = train_patches, y = value, colour = metric)) +
    geom_point(size = 3, alpha = 0.4) +
    geom_line(data = pc_macro_mean, linewidth = 1.2) +
    geom_point(data = pc_macro_mean, size = 4) +
    facet_grid(. ~ mode) +
    scale_colour_manual(values = c("Macro F1" = "#388C73", "Mean IoU" = "#6a4836",
                                   "Overall accuracy" = "grey40"), name = NULL) +
    labs(x = "Realized training patches", y = NULL) +
    theme_param
```


### Does model architecture matter?

- Three arms on `fld_chmret_leafoff`: U-Net (base grid), UNet3+, and the multi-branch
  fusion encoder. Same seeds -> same test patches, and bf64/d5 held constant, so the
  only thing that varies is the architecture.
- Top: model-level Macro F1; middle: per-class F1 coloured by class; bottom: the
  FSW<->UPL confusion rates the fusion encoder is meant to attack.
- Boxplot with seed points overlaid to show the distribution of model runs.
- Reads `arch_compare_long.csv` (tidy, one row per arch x seed), so adding a fourth
  arm to `dl_08b_aggregate_patchcurve.py --arch-dir` needs no change here beyond a
  colour. Arms absent from the CSV are dropped silently.

```{r}
#| label: q7-arch-data

# Display names + palette, keyed by the `arch` values dl_08b writes. Extend both
# to add an arm; everything downstream keys off ARCH_LEVELS.
arch_name  <- c(unet = "U-Net", unet3plus = "UNet3+", mbfusion = "Fusion")
arch_color <- c("U-Net" = "#4477AA", "UNet3+" = "#EE7733", "Fusion" = "#009988")

arch_root <- function(mode_dir) file.path(arch_dir_base, mode_dir, "analysis")

read_arch_csv <- function(file) {
  bind_rows(
    read.csv(file.path(arch_root("multiclass"), file)) |> mutate(mode = "Multiclass"),
    read.csv(file.path(arch_root("binary"), file))     |> mutate(mode = "Binary")
  ) |>
    mutate(mode = factor(mode, levels = c("Multiclass", "Binary")))
}

arch_long <- read_arch_csv("arch_compare_long.csv") |>
  mutate(arch = factor(arch_name[arch], levels = arch_name))

# Levels actually present, so a two-arm (pre-fusion) tree still renders
ARCH_LEVELS <- levels(droplevels(arch_long$arch))
arch_long <- arch_long |> mutate(arch = factor(as.character(arch), levels = ARCH_LEVELS))

# Long-form metrics: macro/mean-IoU, per-class IoU & F1, and confusion rates all
# arrive as columns; melt everything numeric and label the pieces.
arch_metrics <- arch_long |>
  select(-any_of(c("params", "gflops", "sec_per_epoch", "epochs_run", "fit_seconds"))) |>
  pivot_longer(where(is.numeric) & !any_of("seed"),
               names_to = "metric", values_to = "value") |>
  filter(!is.na(value))

arch_class <- arch_metrics |>
  filter(str_detect(metric, "^(iou|f1)_")) |>
  separate_wider_regex(metric, c(metric = "iou|f1", "_", class = ".*")) |>
  mutate(class = factor(class, levels = class_order))
```

```{r}
#| label: q7-arch
#| fig-width: 16
#| fig-height: 16

# Alpha separates arms inside a class group; recycled to however many arms exist.
arch_alpha <- setNames(seq(1, 0.4, length.out = length(ARCH_LEVELS)), ARCH_LEVELS)

arch_fig <- function(m) {
  macro <- arch_metrics |>
    filter(metric == "macro_f1", mode == m) |>
    ggplot(aes(x = arch, y = value, fill = arch)) +
      box_jitter(point_size = 4) +
      scale_fill_manual(values = arch_color, name = NULL) +
      labs(x = NULL, y = "Macro F1", title = paste0(m, ": model-level Macro F1")) +
      theme_param

  cls <- arch_class |>
    filter(metric == "f1", mode == m) |>
    ggplot(aes(x = class, y = value, fill = class, alpha = arch,
               group = interaction(class, arch))) +
      box_jitter(point_size = 4) +
      scale_fill_manual(values = class_color, guide = "none") +
      scale_alpha_manual(values = arch_alpha, name = NULL) +
      guides(alpha = guide_legend(override.aes = list(fill = "grey30"))) +
      labs(x = NULL, y = "F1",
           title = paste0("Per-class F1 (", paste(ARCH_LEVELS, collapse = " vs "), ")")) +
      theme_param

  parts <- list(macro, cls)
  heights <- c(1, 1.4)

  # Confusion panel: multiclass only -- binary has no FSW, so dl_08b omits the
  # conf_ columns there rather than writing zeros.
  conf <- arch_metrics |> filter(mode == m, str_detect(metric, "^conf_.*_as_"))
  if (nrow(conf) > 0) {
    conf_p <- conf |>
      mutate(metric = recode(metric,
                             conf_FSW_as_UPL = "True FSW -> predicted UPL",
                             conf_UPL_as_FSW = "True UPL -> predicted FSW")) |>
      ggplot(aes(x = arch, y = value, fill = arch)) +
        box_jitter(point_size = 4) +
        facet_wrap(~ metric, scales = "free_y") +
        scale_fill_manual(values = arch_color, guide = "none") +
        scale_y_continuous(labels = scales::label_percent(accuracy = 0.1)) +
        labs(x = NULL, y = "Share of the true class's pixels",
             title = "FSW <-> UPL confusion (lower is better)") +
        theme_param
    parts <- c(parts, list(conf_p))
    heights <- c(heights, 1)
  }

  wrap_plots(parts, ncol = 1) + plot_layout(heights = heights)
}

arch_fig("Multiclass")
arch_fig("Binary")
```

Paired per-seed contrasts against the baseline arm. Same seed => same test patches,
so each row is a paired difference, and at n=5 the credible summary is the sign
consistency (`n_better` / `n_seeds`), not a p-value. `lower_is_better` marks the
confusion rates, where a negative delta is the improvement.

```{r}
#| label: q7-contrasts

read_arch_csv("arch_contrasts.csv") |>
  filter(metric %in% c("macro_f1", "mean_iou") | str_detect(metric, "^conf_")) |>
  mutate(arch = arch_name[arch], baseline = arch_name[baseline],
         across(c(baseline_mean, arch_mean, delta_mean, delta_sd), \(x) round(x, 4)),
         # pandas writes "True"/"False"; read.csv leaves those as character
         across(c(sign_consistent, lower_is_better), as.logical),
         seeds = paste0(n_better, "/", n_seeds)) |>
  select(mode, arch, baseline, metric, baseline_mean, arch_mean,
         delta_mean, delta_sd, seeds, sign_consistent, lower_is_better) |>
  arrange(mode, arch, metric) |>
  knitr::kable(caption = "Paired per-seed deltas vs the baseline architecture")
```

What the extra accuracy costs. Params come from the training journal (exact);
GFLOPs from Lightning's model summary in `train.log`, so they are blank for cells
whose Lightning version did not print that line. `sec_per_epoch` is fit-only wall
clock and is only recorded for v3 runs onwards -- it is a same-GPU comparison, not
a portable benchmark.

```{r}
#| label: q7-cost

read_arch_csv("arch_cost.csv") |>
  mutate(arch = arch_name[arch],
         params_M = round(params / 1e6, 2),
         params_x = round(params_x_baseline, 3)) |>
  select(mode, arch, params_M, params_x, gflops, sec_per_epoch, epochs_run) |>
  arrange(mode, desc(params_M)) |>
  knitr::kable(caption = "Cost per architecture (seed means)")
```

### What do the fusion gates attend to?

`dl_11_export_gates.py` writes one `.npz` per held-out patch under
`<cell>/seed<k>/gates/`, holding the per-scale softmax gate weights (one channel
per branch, summing to 1 at every pixel) plus a `gate_summary.json` of the means.

**Read these as spatial patterns within a branch, not as a cross-branch ranking.**
A branch's mean gate weight is confounded by its channel width and by what the
projection layer does downstream, so "terrain averages 0.31" is provenance, not
importance. What is interpretable: where within a scene one branch's weight rises
relative to its own average, and whether the gates stay mixed (entropy near
log(n_branch)) rather than collapsing onto one branch.

```{r}
#| label: q8-gates

gate_files <- Sys.glob(file.path(arch_dir_base, "*", "*_mbfusion", "seed*",
                                 "gates", "gate_summary.json"))

if (length(gate_files) == 0) {
  cat("No gate exports found -- run dl_11_export_gates.py and rsync the .npz back.\n")
} else {
  gate_means <- map_dfr(gate_files, function(f) {
    js <- fromJSON(f, simplifyVector = FALSE)
    seed <- str_extract(f, "seed\\d+")
    mode <- if (str_detect(f, "/binary/")) "Binary" else "Multiclass"
    map_dfr(names(js$per_scale), function(sc) {
      tibble(mode = mode, seed = seed, scale = sc,
             branch = names(js$per_scale[[sc]]),
             mean_gate = as.numeric(unlist(js$per_scale[[sc]])))
    })
  }) |>
    mutate(mode  = factor(mode, levels = c("Multiclass", "Binary")),
           scale = factor(scale, levels = unique(scale)))

  gate_means |>
    ggplot(aes(x = scale, y = mean_gate, colour = branch, group = branch)) +
      stat_summary(fun = mean, geom = "line", linewidth = 1) +
      stat_summary(fun = mean, geom = "point", size = 3) +
      geom_jitter(width = 0.08, alpha = 0.35, size = 2) +
      facet_wrap(~ mode) +
      scale_y_continuous(labels = scales::label_percent(accuracy = 1)) +
      labs(x = "Encoder scale (0 = full resolution)", y = "Mean gate weight",
           colour = NULL,
           title = "Fusion gate weight by branch and scale",
           subtitle = "Seed points jittered; compare a branch across scales, not branches to each other") +
      theme_param
}
```
